---
title: "HGSOC_precursor_comparison"
author: "Jared Sipes"
date: "2025-05-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

This document compares the transcripts upregulated by HGSOC EVs to the transcripts elevated in STIC lesions. 

The paper from which I took the STIC lesion data is below:


Multimodal Spatial Profiling Reveals Immune Suppression and Microenvironment Remodeling in Fallopian Tube Precursors to High-Grade Serous Ovarian Carcinoma
  
  First Authors: Tanjina Kade, Jia-Ren Lin, and Clemnes B. Hug.

  Last Authors: Ronny Drapkin and Sandro Santagata

  PMID: 39704522 DOI: 10.1158/2159-8290.CD-24-1366


# 1. Load Packages

Run the script to upload Nanostring packages. If asked to make further installations, type "y" or "a" as needed.

```{r}
# 
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# The following initializes most up to date version of Bioc
# BiocManager::install(version="3.15")

BiocManager::install("NanoStringNCTools")
BiocManager::install("GeomxTools")
BiocManager::install("GeoMxWorkflows")

# Note:
# Needed to install package lme4, numderiv
library(NanoStringNCTools)
library(GeomxTools)
library(GeoMxWorkflows)

if(packageVersion("GeomxTools") < "2.1" & 
   packageVersion("GeoMxWorkflows") >= "1.0.1"){
    stop("GeomxTools and Workflow versions do not match. Please use the same version. 
    This workflow is meant to be used with most current version of packages. 
    If you are using an older version of Bioconductor please reinstall GeoMxWorkflows and use vignette(GeoMxWorkflows) instead")
}

if(packageVersion("GeomxTools") > "2.1" & 
   packageVersion("GeoMxWorkflows") <= "1.0.1"){
    stop("GeomxTools and Workflow versions do not match. 
         Please use the same version, see install instructions above.")
    
    # to remove current package version
        # remove.packages("GeomxTools")
        # remove.packages("GeoMxWorkflows")
    # see install instructions above 
}
```

Other packages used in the Notebook are listed below. 


```{r}

library(knitr)
library(tidyverse)
library(dplyr)
library(ggforce)
library(writexl)
library(janitor)
library(readxl)
library(cowplot)

# library(patchwork)

# required plotting 

library(ggplot2)

library(latex2exp)

# needed for t-test
library(ggpubr)

library(ggrepel)


```





We need three different types of files to create the initial dataset, plus one for when you are doing this for the first time.

1.   DCCs files -- these contain the expression count data and some info about sequencing data from the next gen sequencing platform used.

2.   PKCs -- the probe assay metadata, which describes which gene targets are present in the data, find at the following link: [GeoMx DSP Configuration Files \| NanoString⁤](https://nanostring.com/products/geomx-digital-spatial-profiler/geomx-dsp-configuration-files/)

    (Do not bother to unzip, place in the appropriate file as is.)

3.   Annotation file - this will contain information about the tissue, segment area and nuclei count, and any other info you choose to provide.



Precursor lesion 



# Load the dataset


```{r}
load(file = "lesions_vp_data.Rdata")

lesions_vp_data




```