---
title: "HGSOC_EV_proteomics_similarity"
author: "Jared Sipes"
date: "2025-05-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summmary


The purpose of this dataset is to help understand the similarities betwene the ovarian cancer EV proteins identified in the following dataset:

Trinidad, C.V., Pathak, H.B., Cheng, S. et al. Lineage specific extracellular vesicle-associated protein biomarkers for the early detection of high grade serous ovarian cancer. Sci Rep 13, 18341 (2023). https://doi.org/10.1038/s41598-023-44050-5


I am using processed data from a bioinformatician that quantifies protein intensity. 

# Packages Used

```{r}
# read in .xlsx

library(readxl)
library(tidyverse)


# For Jacard Index

library(pheatmap)


```


# Load the dataset


```{r}

# read in only the Serous Ovarian Cancer Cases

# note that all proteins after row 4720 are bovine contaminating proteins (picked up from the FBS).
# set nmax = 4720 to exclude these

ovc_EV_protein <- read_xlsx(path = paste0(getwd(), "/HGSOC_EV_proteomics/Godwin_LFQ_all_81_Human_and_FBS_Proteins.xlsx"), sheet = "Serous Ovarian Cancer", n_max = 4720)


```


Some of the EV lines contain replicate data. For example: CL_OV_OVCAR2_1, CL_OV_OVCAR2_2, and CL_OV_OVCAR2_3 are biological replicates.

We want to average these three to get only one CL_OV_OVCAR2 value. 

```{r}




list_cell_lines <- c("OVCAR2", "OVCAR3", "OVCAR4", "OVCAR8", "PEO1", "PEO4")

# use mutate to generate the new column

# Notes:
# := is an assignment argument like "=", but unlike equals allows name on the left to be computer
# !! - the "bang-bang" operator is "unquoting" this tells R to evaluate the variable and use its value

ovc_EV_protein_avg <- ovc_EV_protein

for(cell_line in list_cell_lines){
  # create the new column name
  new_col <- paste0("CL_OV_", cell_line)
  
  # match all col names containing cell line, and take rowmean 
  # .keep = "unused" ensures that the cols used to create the average are removed
  ovc_EV_protein_avg <- ovc_EV_protein_avg |> mutate(!!new_col:= rowMeans(across(matches(cell_line)), na.rm = T), .keep = "unused") 
  
}


```

```{r}

# test <- ovc_EV_protein_avg %>%
#   filter(!is.na(TE_OV_600071)) %>%
#   pull(Accession)  # or select(gene_id)
# 
# 
# test2 <- df[df$name == "TE_OV_600071", ]$list_proteins


```


```{r}

# subset df to get only items of interest

EV_data <- select(ovc_EV_protein_avg, matches("_OV"))

name_EV <- names(EV_data)



df <- tibble(name = name_EV, num_proteins= numeric(length(name_EV)), list_proteins = list(length(name_EV)))

for(name in name_EV){
  df[df$name == name, ]$num_proteins <- length(which(EV_data[name] >0))
  df[df$name == name, ]$list_proteins <-  list(ovc_EV_protein_avg$Accession[which(EV_data[name] >0)])
  
}


rownames(df) <- df$name

```





Ideas for displaying:

Venn won't work well - there are in fact 6-venn diagrams, for example here: https://www.interactivenn.net/index.html, but good luck reading one! 




Pairwise Jaccard Similarity Heatmap

Upset plot

# Jacard Similarity Plot


```{r}


protein_lists <- df$list_proteins

names(protein_lists) <- df$name

# Function to compute Jaccard index
jaccard_index <- function(x, y) {
  length(intersect(x, y)) / length(union(x, y))
}

# Compute pairwise Jaccard matrix
samples <- names(protein_lists)
jaccard_matrix <- outer(samples, samples, Vectorize(function(a, b) jaccard_index(protein_lists[[a]], protein_lists[[b]])))
rownames(jaccard_matrix) <- colnames(jaccard_matrix) <- samples

# Plot heatmap
library(pheatmap)
pheatmap(jaccard_matrix, display_numbers = TRUE, main = "Jaccard Similarity Between Samples")


```







Bar plot: percent shared by all, percent shared by 5, by 4, 3, 2, 1


```{r}

all_samples <- df
CL_samples <- df[grepl("CL", df$name), ]

add_hits <- function(df){
    df <- df
    protein_hits <- list()
  
  # loop from 1: 15
    
  for(i in 1:length(df$list_proteins)){
      # get the list of proteins
      protein_list <- df$list_proteins[[i]]
      # creat a new empty list of hits
      hits <- c()
    for(protein in protein_list){
      # calculate number of hits
      num_hits <- sapply(df$list_proteins, function(x) protein %in% x) |> sum()
      #append to list
      hits <- c(hits, num_hits)
    }
    protein_hits[[i]] <- hits
  }
  
   df$protein_hits <- protein_hits
   return(df)
}

all_samples <- add_hits(all_samples)
CL_samples <- add_hits(CL_samples)

```



```{r}


# takes the df and creates a new df with a column for number and source (cell line or tissue explant)

collapse_df <- function(df){
  all_data <- data.frame()
  for(name in df$name){
    list <- as.vector(df[df$name == name, ]$protein_hits[[1]])
    df_new <- data.frame(number = list, source = name)
    all_data <- bind_rows(all_data, df_new)
  }
  return(all_data)
}


all_data <- collapse_df(all_samples)
CL_data <- collapse_df(CL_samples)




#only_CL <- all_data[grepl("CL", all_data$source), ]

# Plot
ggplot(CL_data, aes(x=source, fill = factor(number))) +
  geom_bar(position = "stack") +
  labs(title = "Frequently occuring proteins in HGSOC-EVs",
       x = "Sample Origin",
       y = "Number of Proteins",
       fill = "Protein found in \n n samples") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle=90, vjust=.5, hjust=1))
  

```


