---
title: "HGSOC_precursor_comparison"
author: "Jared Sipes"
date: "2025-05-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

This document compares the transcripts upregulated by HGSOC EVs to the transcripts elevated in STIC lesions. 

The paper from which I took the STIC lesion data is below:


Multimodal Spatial Profiling Reveals Immune Suppression and Microenvironment Remodeling in Fallopian Tube Precursors to High-Grade Serous Ovarian Carcinoma
  
  First Authors: Tanjina Kade, Jia-Ren Lin, and Clemnes B. Hug.

  Last Authors: Ronny Drapkin and Sandro Santagata

  PMID: 39704522 DOI: 10.1158/2159-8290.CD-24-1366

Notes:

The .dcc files for the negetaive control well are not included in the online repository. 

data_input/annotation/Annotations_FTE_Reprogramming_4-24-25.xlsx
precursor_data/annotation/GSE281193_annotation.xlsx
precursor_data/annotation/annotation_corrected_4-28-25.xlsx
precursor_data/annotation/cd-24-1366_supplementary_table_s2_suppst2.xlsx
# 1. Load Packages

Run the script to upload Nanostring packages. If asked to make further installations, type "y" or "a" as needed.

```{r}
# 
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# The following initializes most up to date version of Bioc
BiocManager::install()

BiocManager::install("NanoStringNCTools")
BiocManager::install("GeomxTools")
BiocManager::install("GeoMxWorkflows")

# Note:
# Needed to install package lme4, numderiv
library(NanoStringNCTools)
library(GeomxTools)
library(GeoMxWorkflows)

if(packageVersion("GeomxTools") < "2.1" & 
   packageVersion("GeoMxWorkflows") >= "1.0.1"){
    stop("GeomxTools and Workflow versions do not match. Please use the same version. 
    This workflow is meant to be used with most current version of packages. 
    If you are using an older version of Bioconductor please reinstall GeoMxWorkflows and use vignette(GeoMxWorkflows) instead")
}

if(packageVersion("GeomxTools") > "2.1" & 
   packageVersion("GeoMxWorkflows") <= "1.0.1"){
    stop("GeomxTools and Workflow versions do not match. 
         Please use the same version, see install instructions above.")
    
    # to remove current package version
        # remove.packages("GeomxTools")
        # remove.packages("GeoMxWorkflows")
    # see install instructions above 
}
```

Other packages used in the Notebook are listed below. 


```{r}

library(knitr)
library(tidyverse)
library(dplyr)
library(ggforce)
library(writexl)
library(janitor)
library(readxl)
library(cowplot)

# library(patchwork)

# required plotting 

library(ggplot2)

library(latex2exp)

# needed for t-test
library(ggpubr)

library(ggrepel)


```


# 2. Create GeoMx Object 


We need three different types of files to create the initial dataset, plus one for when you are doing this for the first time.

1.   DCCs files -- these contain the expression count data and some info about sequencing data from the next gen sequencing platform used.

2.   PKCs -- the probe assay metadata, which describes which gene targets are present in the data, find at the following link: [GeoMx DSP Configuration Files \| NanoString⁤](https://nanostring.com/products/geomx-digital-spatial-profiler/geomx-dsp-configuration-files/)

    (Do not bother to unzip, place in the appropriate file as is.)

3.   Annotation file - this will contain information about the tissue, segment area and nuclei count, and any other info you choose to provide.



Precursor lesion transcriptome data is found in `/precursor_data` and subfolders


```{r}

# Reference the main folder 'file.path' containing the sub-folders with each
# data file type:

datadir <- file.path(paste0(getwd(), "/precursor_data"))

# automatically list files in each directory for use
DCCFiles <- dir(file.path(datadir, "dccs"), pattern = "*.dcc",
                full.names = TRUE, recursive = TRUE)


PKCFiles <- dir(file.path(datadir, "pkcs"), pattern = ".pkc$",
                                full.names = TRUE, recursive = TRUE)


# Make sure to add correct annotation file
SampleAnnotationFile <- 
    dir(file.path(datadir, "annotation"), pattern = "4-28-25.xlsx$",
        full.names = TRUE, recursive = TRUE)


```



Testing creation of `readNanoStringGeoMxSet` object. If you get errors saying files are missing or do not have any count info, check to make sure all .dcc files are in the folder (and that you have copied the correct ones).

```{r}


#load data

allGeoMx <-
    readNanoStringGeoMxSet(dccFiles = DCCFiles,
                           pkcFiles = PKCFiles,
                           phenoDataFile = SampleAnnotationFile,
                           phenoDataSheet = "Sheet1", # make sure this matches doc
                           phenoDataDccColName = "Sample_ID",
                           protocolDataColNames = c("Aoi", "Roi"),
                           experimentDataColNames = c("Panel"))

# Potential Errors:
# Error in .subset2(x, i, exact = exact) : subscript out of bounds
# Check Traceback
# If traceback goes to pheno[[nucleiCol]]
# check the naming of nuclei columns -- should have only one! 

```

# 2.2 Subsetting Dataset

```{r}
epithelial <- allGeoMx[, pData(allGeoMx)$Cell_Types == "epithelial"]

stroma <- allGeoMx[, pData(allGeoMx)$Days == "stroma"]
```




# 4. QC and Pre-processing

There are three sets of pre-process workflow for GeoMx data. In the first part, faulty segments are removed and genes are selected based on Quality Control.

First, however, we must shift all counts of 0 to 1 (this permits downstream transformations.

```{r}


# shift counts to one

allGeoMx <- shiftCountsOne(allGeoMx, useDALogic = TRUE)

```




## 4.1 Select Segment QC

First, we select the QC parameter cutoffs, against which our ROI/AOI segments will be tested and flagged appropriately. We have selected the appropriate study-specific parameters for this study. Note: the default QC values recommended above are advised when surveying a new dataset for the first time.

For some reason the QC Summary seems not to be working properly for low area? Possibly the area got dropped from the annotations somehow? But areas seems to be there based on a check?


```{r}

# Default QC cutoffs are commented in () adjacent to the respective parameters
# study-specific values were selected after visualizing the QC results in more
# detail below

QC_params <-
    list(minSegmentReads = 1000, # Minimum number of reads (1000)
         percentTrimmed = 80,    # Minimum % of reads trimmed (80%)
         percentStitched = 80,   # Minimum % of reads stitched (80%)
         percentAligned = 80,    # Minimum % of reads aligned (80%)
         percentSaturation = 50, # Minimum sequencing saturation (50%)
         minNegativeCount = 1,   # Minimum negative control counts (10)
         maxNTCCount = 1000,     # Maximum counts observed in NTC well (1000)
         minNuclei = 20,         # Minimum # of nuclei estimated (100)
         minArea = 1000)         # Minimum segment area (5000)
allGeoMx <-
    setSegmentQCFlags(allGeoMx, 
                      qcCutoffs = QC_params)        

# Collate QC Results
QCResults <- protocolData(allGeoMx)[["QCFlags"]]
  # this seems to generate some NA values for the NTc controls, which don't have area. 
  # Tell it to ignore NAs? Or drop NTC controls?


flag_columns <- colnames(QCResults)
QC_Summary <- data.frame(Pass = colSums(!QCResults[, flag_columns], na.rm = TRUE),
                         Warning = colSums(QCResults[, flag_columns], na.rm = TRUE))
QCResults$QCStatus <- apply(QCResults, 1L, function(x) {
    ifelse(sum(x) == 0L, "PASS", "WARNING")
})
QC_Summary["TOTAL FLAGS", ] <-
    c(sum(QCResults[, "QCStatus"] == "PASS", na.rm = TRUE),
      sum(QCResults[, "QCStatus"] == "WARNING", na.rm = TRUE))

```


```{r}

# make sure you loaded knitr using library(knitr)

pkcs <- annotation(allGeoMx)
modules <- gsub(".pkc", "", pkcs)
kable(data.frame(PKCs = pkcs, modules = modules))

```



## 4.2 Visualize Segment QC

Before excluding any low-performing ROI/AOI segments, we visualize the distributions of the data for the different QC parameters. Note that the "Select Segment QC" and "Visualize Segment QC" sections are performed in parallel to fully understand low-performing segments for a given study. Iteration may follow to select the study-specific QC cutoffs.

For QC visualization, we write a quick function to draw histograms of our data.

```{r}



# Graphical summaries of QC statistics plot function
QC_histogram <- function(assay_data = NULL,
                         annotation = NULL,
                         fill_by = NULL,
                         thr = NULL,
                         scale_trans = NULL) {
    plt <- ggplot(assay_data,
                  aes_string(x = paste0("unlist(`", annotation, "`)"),
                             fill = fill_by)) +
        geom_histogram(bins = 50) +
        geom_vline(xintercept = thr, lty = "dashed", color = "black") +
        theme_bw() + guides(fill = "none") +
        facet_wrap(as.formula(paste("~", fill_by)), nrow = 4) +
        labs(x = annotation, y = "Number of Segments", title = annotation)
    if(!is.null(scale_trans)) {
        plt <- plt +
            scale_x_continuous(trans = scale_trans)
    }
    plt
}

```


Now we explore each of the QC metrics for the segments.




```{r fig.height=5, fig.width=8}

col_by <- "Cell_Types"

trimmed <- QC_histogram(sData(allGeoMx), "Trimmed (%)", col_by, 80)
stitched <- QC_histogram(sData(allGeoMx), "Stitched (%)", col_by, 5)
aligned <- QC_histogram(sData(allGeoMx), "Aligned (%)", col_by, 5)
saturated <- QC_histogram(sData(allGeoMx), "Saturated (%)", col_by, 50) +
    labs(title = "Sequencing Saturation (%)",
         x = "Sequencing Saturation (%)")
#AreaHist <- QC_histogram(sData(allGeoMx), "area", col_by, 1000, scale_trans = "log10")
#NucleiHist <- QC_histogram(sData(allGeoMx), "nuclei", col_by, 20)

QC_metrics_by_day <- plot_grid(trimmed, stitched, aligned, saturated)

QC_metrics_by_day
```



```{r}
# calculate the negative geometric means for each module
negativeGeoMeans <- 
    esBy(negativeControlSubset(allGeoMx), 
         GROUP = "Module", 
         FUN = function(x) { 
             assayDataApply(x, MARGIN = 2, FUN = ngeoMean, elt = "exprs") 
         }) 
protocolData(allGeoMx)[["NegGeoMean"]] <- negativeGeoMeans

# explicitly copy the Negative geoMeans from sData to pData
negCols <- paste0("NegGeoMean_", modules)
pData(allGeoMx)[, negCols] <- sData(allGeoMx)[["NegGeoMean"]]
for(ann in negCols) {
    plt <- QC_histogram(pData(allGeoMx), ann, col_by, 2, scale_trans = "log10")
    print(plt)
}

```




```{r}
# detach neg_geomean columns ahead of aggregateCounts call
pData(allGeoMx) <- pData(allGeoMx)[, !colnames(pData(allGeoMx)) %in% negCols]

# show all NTC values, Freq = # of Segments with a given NTC count:

# we seem to be missing a column called NTC that is required for this to work 
# kable(table(NTC_Count = sData(allGeoMx)$NTC),
#       col.names = c("NTC Count", "# of Segments"))

```



```{r}

kable(QC_Summary, caption = "QC Summary Table for each Segment")


```





## 4.3 Remove Flagged Segments

As the final step in the QC, we remove all flagged segments that do not meet the QC cutoff.

```{r}
allGeoMx <- allGeoMx[, QCResults$QCStatus == "PASS"]

# Subsetting our dataset has removed samples which did not pass QC

dim(allGeoMx)


```




```{r}


# Generally keep the qcCutoffs parameters unchanged. Set removeLocalOutliers to 
# FALSE if you do not want to remove local outliers
allGeoMx <- setBioProbeQCFlags(allGeoMx, 
                               qcCutoffs = list(minProbeRatio = 0.1,
                                                percentFailGrubbs = 20), 
                               removeLocalOutliers = TRUE)

ProbeQCResults <- fData(allGeoMx)[["QCFlags"]]

# Define QC table for Probe QC
qc_df <- data.frame(Passed = sum(rowSums(ProbeQCResults[, -1]) == 0),
                    Global = sum(ProbeQCResults$GlobalGrubbsOutlier),
                    Local = sum(rowSums(ProbeQCResults[, -2:-1]) > 0
                                & !ProbeQCResults$GlobalGrubbsOutlier))

```


We report the number of global and local outlier probes.


```{r}

qc_df


```



## 5.2 Exclude Outlier Probes

```{r}
#Subset object to exclude all that did not pass Ratio & Global testing
ProbeQCPassed <- 
    subset(allGeoMx, 
           fData(allGeoMx)[["QCFlags"]][,c("LowProbeRatio")] == FALSE &
               fData(allGeoMx)[["QCFlags"]][,c("GlobalGrubbsOutlier")] == FALSE)
dim(ProbeQCPassed)
#> Features  Samples 
#>    18641      229
allGeoMx <- ProbeQCPassed 


```



## 5.3 Create Gene Level Count Data

With our Probe QC steps complete, we will generate a gene-level count matrix. The count for any gene with multiple probes per segment is calculated as the geometric mean of those probes.

```{r}

# Check how many unique targets the object has
length(unique(featureData(allGeoMx)[["TargetName"]]))
#> [1] 18504

# collapse to targets
target_allGeoMx <- aggregateCounts(allGeoMx)
dim(target_allGeoMx)
#> Features  Samples 
#>    18504      229
exprs(target_allGeoMx)[1:5, 1:2]
#>       DSP-1001250007851-H-A02.dcc DSP-1001250007851-H-A03.dcc
#> A2M                           485                         262
#> NAT2                           15                          18
#> ACADM                          31                          15
#> ACADS                          27                          17
#> ACAT1                          29                          24


```


## 5.4 Limit of Quantitation

In addition to Segment and Probe QC, we also determine the limit of quantification (LOQ) per segment. The LOQ is calculated based on the distribution of negative control probes and is intended to approximate the quantifiable limit of gene expression per segment. Please note that this process is more stable in larger segments. Likewise, the LOQ may not be as accurately reflective of true signal detection rates in segments with low negative probe counts (ex: \<2). The formula for calculating the LOQ in the ith segment is:

$$LOQ_i=geomean(NegProbe_i)∗geoSD(NegProbe_i)^n$$

We typically use 2 geometric standard deviations (n=2) above the geometric mean as the LOQ, which is reasonable for most studies. We also recommend that a minimum LOQ of 2 be used if the LOQ calculated in a segment is below this threshold.


```{r}

# Define LOQ SD threshold and minimum value
cutoff <- 2
minLOQ <- 2

# Calculate LOQ per module tested
LOQ <- data.frame(row.names = colnames(target_allGeoMx))
for(module in modules) {
    vars <- paste0(c("NegGeoMean_", "NegGeoSD_"),
                   module)
    if(all(vars[1:2] %in% colnames(pData(target_allGeoMx)))) {
        LOQ[, module] <-
            pmax(minLOQ,
                 pData(target_allGeoMx)[, vars[1]] * 
                     pData(target_allGeoMx)[, vars[2]] ^ cutoff)
    }
}
pData(target_allGeoMx)$LOQ <- LOQ



```



## 5.5 Filtering

After determining the limit of quantification (LOQ) per segment, we recommend filtering out either segments and/or genes with abnormally low signal. Filtering is an important step to focus on the true biological data of interest.

We determine the number of genes detected in each segment across the dataset.

```{r}


LOQ_Mat <- c()
for(module in modules) {
    ind <- fData(target_allGeoMx)$Module == module
    Mat_i <- t(esApply(target_allGeoMx[ind, ], MARGIN = 1,
                       FUN = function(x) {
                           x > LOQ[, module]
                       }))
    LOQ_Mat <- rbind(LOQ_Mat, Mat_i)
}
# ensure ordering since this is stored outside of the geomxSet
LOQ_Mat <- LOQ_Mat[fData(target_allGeoMx)$TargetName, ]


```


## 5.6 Segment Gene Detection

We first filter out segments with exceptionally low signal. These segments will have a small fraction of panel genes detected above the LOQ relative to the other segments in the study. Let's visualize the distribution of segments with respect to their % genes detected:

```{r fig.height=10, fig.width=20}

library(patchwork)

# Save detection rate information to pheno data
pData(target_allGeoMx)$GenesDetected <- 
    colSums(LOQ_Mat, na.rm = TRUE)
pData(target_allGeoMx)$GeneDetectionRate <-
    pData(target_allGeoMx)$GenesDetected / nrow(target_allGeoMx)

# Determine detection thresholds: 1%, 5%, 10%, 15%, >15%
pData(target_allGeoMx)$DetectionThreshold <- 
    cut(pData(target_allGeoMx)$GeneDetectionRate,
        breaks = c(0, 0.01, 0.05, 0.1, 0.15, 0.20, 0.25, 1),
        labels = c("<1%", "1-5%", "5-10%", "10-15%", "16%-20%", "21%-25%", ">25%"))

# stacked bar plot of different cut points (1%, 5%, 10%, 15%)
a<- ggplot(pData(target_allGeoMx),
       aes(x = DetectionThreshold)) +
    geom_bar(aes(fill = Lesions_Types)) +
    geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
    theme_bw() +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    labs(x = "Gene Detection Rate",
         y = "Segments, #",
         fill = "Segment Type")
#> Warning: The dot-dot notation (`..count..`) was deprecated in ggplot2 3.4.0.
#> ℹ Please use `after_stat(count)` instead.
#> This warning is displayed once every 8 hours.
#> Call `lifecycle::last_lifecycle_warnings()` to see where this warning was
#> generated.


b<- ggplot(pData(target_allGeoMx),
       aes(x = DetectionThreshold)) +
    geom_bar(aes(fill = BRCA_status)) +
    geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
    theme_bw() +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    labs(x = "Gene Detection Rate",
         y = "Segments, #",
         fill = "Segment Type")


c <- ggplot(pData(target_allGeoMx),
       aes(x = DetectionThreshold)) +
    geom_bar(aes(fill = PatientID)) +
    geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
    theme_bw() +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    labs(x = "Gene Detection Rate",
         y = "Segments, #",
         fill = "Segment Type")

d <- ggplot(pData(target_allGeoMx),
       aes(x = DetectionThreshold)) +
    geom_bar(aes(fill = Cell_Types)) +
    geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
    theme_bw() +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    labs(x = "Gene Detection Rate",
         y = "Segments, #",
         fill = "Segment Type")


a + b+ c + d

```


Comments: It looks like stroma has a higher chance of having a low gene detection rate than epithelia. Potentially due to lower gene expression in stroma? Or fewer cells per ROI?




```{r}

# cut percent genes detected from above
kable(table(pData(target_allGeoMx)$DetectionThreshold,
            pData(target_allGeoMx)$Cell_Types))

kable(table(pData(target_allGeoMx)$DetectionThreshold,
            pData(target_allGeoMx)$PatientID))


kable(table(pData(target_allGeoMx)$DetectionThreshold,
            pData(target_allGeoMx)$Lesions_Types))

```




In this example, we choose to remove segments with less than 5% of the genes detected. Generally, 5-10% detection is a reasonable segment filtering threshold. However, based on the experimental design (e.g. segment types, size, nuclei) and tissue characteristics (e.g. type, age), these guidelines may require adjustment.

```{r}
target_allGeoMx <-
    target_allGeoMx[, pData(target_allGeoMx)$GeneDetectionRate >= .05]

# from the paper, the 5% cutoff was used

dim(target_allGeoMx)
#> Features  Samples 
#>    18677    566

```



We seem to have an issue where one segment is not passing QC that passed in the original dataset. (566 in this dataset vs 567 reported in the paper.)

It's sequencing saturation is about 40, which implies it should have been removed. Based on this, it appears that the statement that segments were removed with "<50% sequencing saturation" is incorrect ?? They musthave used <40%??

This is unlikely to cause very significant effects on downstream calculations, so I will proceed. 



### 4.5.2 Gene Detection Rate.

Next, we determine the detection rate for genes across the study. To illustrate this idea, we create a small gene list (goi) to review.

```{r}
library(scales) # for percent

# Calculate detection rate:
LOQ_Mat <- LOQ_Mat[, colnames(target_allGeoMx)]
fData(target_allGeoMx)$DetectedSegments <- rowSums(LOQ_Mat, na.rm = TRUE)
fData(target_allGeoMx)$DetectionRate <-
    fData(target_allGeoMx)$DetectedSegments / nrow(pData(target_allGeoMx))

# Gene of interest detection table
goi <- c("KRT18", "CCL2",
         "FOXJ1", "PAX8", "LGR5")
goi_df <- data.frame(
    Gene = goi,
    Number = fData(target_allGeoMx)[goi, "DetectedSegments"],
    DetectionRate = percent(fData(target_allGeoMx)[goi, "DetectionRate"]))



```

```{r}

goi_df


```





# Load the dataset


```{r}
load(file = "lesions_vp_data.Rdata")

lesions_vp_data




```