---
title: "main_figures"
author: "Jared Sipes"
date: "2025-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary 

Contains code used to generate main figures.

Note that Figures 1 and 2 do not contain any figures generated using the GeoMx or Mass spec data, and are therefore not included.

Supplemental_Figures:

Supplemental Figures 3, 10, and 12-16 are found in this document. 
Supplemental Figure 1 uses data from another study, and may be found in HGSOC_EV_proteomics.
Supplemental Figure 2 may be found in QC_and_normalization. 
Supplemental Figure 11 uses data from another study, and may be found in HGSOC_precursor_comparison.Rmd


All either supplemental figures were either generated with GraphPad prism or do not contain plots, with the exception of Supplemental Figure 2, which may be found in the file QC_and_normalization.Rmd. 



## 1. Packages Needed


```{r}
# #
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# # 
# # # The following initializes most up to date version of Bioc
# BiocManager::install()
# 
# BiocManager::install("NanoStringNCTools")
# BiocManager::install("GeomxTools")
# BiocManager::install("GeoMxWorkflows")

# Note:
# Needed to install package lme4, numderiv
library(NanoStringNCTools)
library(GeomxTools)
library(GeoMxWorkflows)

if(packageVersion("GeomxTools") < "2.1" & 
   packageVersion("GeoMxWorkflows") >= "1.0.1"){
    stop("GeomxTools and Workflow versions do not match. Please use the same version. 
    This workflow is meant to be used with most current version of packages. 
    If you are using an older version of Bioconductor please reinstall GeoMxWorkflows and use vignette(GeoMxWorkflows) instead")
}

if(packageVersion("GeomxTools") > "2.1" & 
   packageVersion("GeoMxWorkflows") <= "1.0.1"){
    stop("GeomxTools and Workflow versions do not match. 
         Please use the same version, see install instructions above.")
    
    # to remove current package version
        # remove.packages("GeomxTools")
        # remove.packages("GeoMxWorkflows")
    # see install instructions above 
}
```

```{r}


# Used to read in Excel Documents

library(readxl)

# general tools for manipulating data

library(dplyr)

#Needed for volcano plotting functions

library(ggrepel)
library(ggplot2)
library(latex2exp)

#Go_plots

library(stringr) # need for string_wrap


# Combining Graphs

library(patchwork)

# Venn Diagram 


library(ggvenn)
library(eulerr)
library(ggpubr)


# Heatmap

library(ComplexHeatmap)

library(grid)  # For capturing the heatmap as a grob

# for a color palette

library(wesanderson)
library(khroma)
library(circlize)

```



## 2. Load Files

#### 2.1  Load GeoMx Object

The GeoMx Object was created using the QC_and_normalization.Rmd file. 
It is not separately uploaded in the Github repository due to size and must be generated using the QC_and_normalization.Rmd file. 

```{r}

# Create a new column called `Patient` that contains the new order of patient numbers. 
# Only day1 tissue is renamed for now. 


pData(EV_q_norm) <- pData(EV_q_norm) |> mutate(Patient = recode(Tissue,
                                                 "461-27304" = "pt1",
                                                 "5929-606950" = "pt2",
                                                 "5929-607267" = "pt3",
                                                 "5929-606994" = "pt4",
                                                 "5929-607140" = "pt5",
                                                 "NW71422" = "pt6",
                                                 "461-27413" = "pt7",
                                                 "UH00006" = "pt8",
                                                 "UH000025" = "pt9",
                                                 "5929-606939" = "pt10",
                                                 "461-27444" = "pt11",
                                                 "5929-606856" = "pt12",
                                                 "UH000028" = "pt13",
                                                 "5929-608036" = "pt14",
                                                 "021355" = "pt15",
                                                 "5929-607836" = "pt16",
                                                 "UH00022" = "pt17",
                                                 "461-27355" = "pt18"
                                                 ))

```


## 2. New T-test


```{r}


t.test_logfold <- function(obj, variable_type = "Treatment", upreg_group = "Ciliated", downreg_group = "Secretory", elt_name = "q_norm"){
  
  # Get locations of all in the "up" group for that variable
  g_up <- which(pData(obj)[variable_type] == upreg_group)
  # get locations of all in the "down" group for that variable
  g_down <- which(pData(obj)[variable_type] == downreg_group)
  
  # Retrieve normalized expression data from the GeoMxDataSet Object
  obj <-  assayDataElement(object = obj, elt = elt_name)
  
  # For each gene (every row), perform a T-test comparing the columns in the 
  #up_group vs down_group
  
  t.test <- apply(obj, 1, 
              function(x) t.test(
                x[g_up],
                x[g_down]))
  
  #calculate the log_2 Fold Change of up_group compared to down for each row
  
  log_fold <- apply(obj, 1,
                    function(x)  log2(mean(x[g_up])/mean(x[g_down]))
                    )
  
  # convert to data frame
  log_fold <- as.data.frame(log_fold)
  
  # retrieve the p.values from the t.test object and store to a datafram

  df <- do.call(rbind, lapply(t.test, function(x) c(
    p_value = unname(x$p.value))))
  
  # create an adjusted p_value using the Bonferooni correction 

  df <- transform(df, padj = p.adjust(p_value, method = "BH"))
  
  # merge the log_fold values with the p_values
  
  output <- merge(df, log_fold, by = 0)
  
  
  # rename the column rownames to Gene
  
  output <- output |> dplyr::rename(Marker.name = Row.names, p_value_t = p_value)
  #  Add the -log10 of the pvalue for plotting 
  output <- output |> mutate(minus_log10.p.value_t = -log10(p_value_t))
  
  return(output)
  
}


```


# Separate into Day1 and Day14 Cil and Sec

```{r}
dataEV_D14 <- EV_q_norm[, pData(EV_q_norm)$Days == "14-Day"]

dataEV_D14_sec <- dataEV_D14[, pData(dataEV_D14)$segment == "Secretory"]
dataEV_D14_cil <- dataEV_D14[, pData(dataEV_D14)$segment == "Ciliated"]

dataEV_D01 <- EV_q_norm[, pData(EV_q_norm)$Days == "1-Day"]

dataEV_D01_sec <- dataEV_D01[, pData(dataEV_D01)$segment == "Secretory"]
dataEV_D01_cil <- dataEV_D01[, pData(dataEV_D01)$segment == "Ciliated"]

```


```{r}

library(writexl)

# function to save the volcano plot comparisons to excel file. 

save_vp_comparisons <- function(name = "name", PBS.vs.FT240 = NA, FT240.vs.OVCAR3 = NA, PBS.vs.OVCAR3 = NA){
  
    # Create List of data frames
    # format is name = df
    dfs <- list(PBS.vs.FT240 = PBS.vs.FT240, FT240.vs.OVCAR3 = FT240.vs.OVCAR3, PBS.vs.OVCAR3 = PBS.vs.OVCAR3)
    
    # Export
     write_xlsx(dfs, path = paste0(name, ".xlsx"))
     
     print(paste0("File saved to ", getwd(), "/", name, ".xlsx"))
}

# Secretory 1 Day


sec_D1_PBS.vs.FT240 <- t.test_logfold(obj = dataEV_D01_sec, variable_type = "Treatment", upreg = "FT240", downreg = "PBS")
sec_D1_FT240.vs.OVCAR3 <- t.test_logfold(obj = dataEV_D01_sec, variable_type = "Treatment", upreg = "OVCAR3", downreg = "FT240")
sec_D1_PBS.vs.OVCAR3 <- t.test_logfold(obj = dataEV_D01_sec, variable_type = "Treatment", upreg = "OVCAR3", downreg = "PBS")

save_vp_comparisons(name = "1Day_Secretory_Comparison_ttest", PBS.vs.FT240 = sec_D1_PBS.vs.FT240, FT240.vs.OVCAR3 = sec_D1_FT240.vs.OVCAR3, PBS.vs.OVCAR3 = sec_D1_PBS.vs.OVCAR3)


# Ciliated 1 Day

cil_D1_PBS.vs.FT240 <- t.test_logfold(obj = dataEV_D01_cil, variable_type = "Treatment", upreg = "FT240", downreg = "PBS")
cil_D1_FT240.vs.OVCAR3 <- t.test_logfold(obj = dataEV_D01_cil, variable_type = "Treatment", upreg = "OVCAR3", downreg = "FT240")
cil_D1_PBS.vs.OVCAR3 <- t.test_logfold(obj = dataEV_D01_cil, variable_type = "Treatment", upreg = "OVCAR3", downreg = "PBS")

save_vp_comparisons(name = "1Day_Ciliated_Comparison_ttest", PBS.vs.FT240 = cil_D1_PBS.vs.FT240, FT240.vs.OVCAR3 = cil_D1_FT240.vs.OVCAR3, PBS.vs.OVCAR3 = cil_D1_PBS.vs.OVCAR3)



# Secretory Day 14

sec_D14_PBS.vs.FT240 <- t.test_logfold(obj = dataEV_D14_sec, variable_type = "Treatment", upreg = "FT240", downreg = "PBS")
sec_D14_FT240.vs.OVCAR3 <- t.test_logfold(obj = dataEV_D14_sec, variable_type = "Treatment", upreg = "OVCAR3", downreg = "FT240")
sec_D14_PBS.vs.OVCAR3 <- t.test_logfold(obj = dataEV_D14_sec, variable_type = "Treatment", upreg = "OVCAR3", downreg = "PBS")

save_vp_comparisons(name = "14Day_Secretory_Comparison_ttest", PBS.vs.FT240 = sec_D14_PBS.vs.FT240, FT240.vs.OVCAR3 = sec_D14_FT240.vs.OVCAR3, PBS.vs.OVCAR3 = sec_D14_PBS.vs.OVCAR3)



# Ciliated Day 14

cil_D14_PBS.vs.FT240 <- t.test_logfold(obj = dataEV_D14_cil, variable_type = "Treatment", upreg = "FT240", downreg = "PBS")
cil_D14_FT240.vs.OVCAR3 <- t.test_logfold(obj = dataEV_D14_cil, variable_type = "Treatment", upreg = "OVCAR3", downreg = "FT240")
cil_D14_PBS.vs.OVCAR3 <- t.test_logfold(obj = dataEV_D14_cil, variable_type = "Treatment", upreg = "OVCAR3", downreg = "PBS")

save_vp_comparisons(name = "14Day_Ciliated_Comparison_ttest", PBS.vs.FT240 = cil_D14_PBS.vs.FT240, FT240.vs.OVCAR3 = cil_D14_FT240.vs.OVCAR3, PBS.vs.OVCAR3 = cil_D14_PBS.vs.OVCAR3)




```

```{r}


get_go_genes <- function(df = sec_D1_PBS.vs.FT240, p_val_cutoff = 0.05, logfc_cutoff = 0.5){
  if(logfc_cutoff > 0){
    gene_list<- df[df$p_value_t < p_val_cutoff & df$log_fold > logfc_cutoff, ]$Marker.name
  }
  if(logfc_cutoff < 0){
    gene_list<- df[df$p_value_t < p_val_cutoff & df$log_fold < logfc_cutoff, ]$Marker.name
    
  }
  return(gene_list)
}




```


```{r}


get_genes <- function(df_FT240 = sec_D1_PBS.vs.FT240, df_OVCAR3 = sec_D1_PBS.vs.OVCAR3){
  genes <- list()
  
  genes$FT240_up <- get_go_genes(df = df_FT240)
  genes$OVCAR3_up <- get_go_genes(df = df_OVCAR3)
  genes$FT240_down <- get_go_genes(df = df_FT240, logfc_cutoff = -0.5)
  genes$OVCAR3_down <- get_go_genes(df = df_OVCAR3, logfc_cutoff = -0.5)
  genes$OVCAR3_unique_up <- setdiff(genes$OVCAR3_up,  genes$FT240_up)
  genes$OVCAR3_unique_down <- setdiff(genes$OVCAR3_down,  genes$FT240_down)
  
  return(genes)
}

# Sec D1

D1_sec <- get_genes(df_FT240 = sec_D1_PBS.vs.FT240, df_OVCAR3 = sec_D1_PBS.vs.OVCAR3)

# Cil D1

D1_cil <- get_genes(df_FT240 = cil_D1_PBS.vs.FT240, df_OVCAR3 = cil_D1_PBS.vs.OVCAR3)

# Sec_D14

D14_sec <- get_genes(df_FT240 = sec_D14_PBS.vs.FT240, df_OVCAR3 = sec_D14_PBS.vs.OVCAR3)

# Cil_D14

D14_cil <- get_genes(df_FT240 = cil_D14_PBS.vs.FT240, df_OVCAR3 = cil_D14_PBS.vs.OVCAR3)


```



#### 2.2  Load GO Files Unique DEGs

These GO analysis files are created using transcripts that were uniquely upregulated ONLY by OVCAR3 EVs, but NOT by FT240 EVs. 

See methods for more details. 


```{r}

# # source of all related .csv files
# source <- paste0(getwd(), "/GO_analysis/unique_OVCAR3_transcripts_only")
# 
# # Read in the .csv files to be used. 
# 
# BP_OVC_sec_new <- read.csv(file = paste0(source, "/sec_GO_BP_OVCAR_upreg.csv"))
# 
# CC_PBS_sec_new <- read.csv(file = paste0(source, "/sec_GO_CC_OVCAR_downreg.csv"))
# 
# BP_OVC_cil_new <- read.csv(file = paste0(source, "/cil_GO_BP_OVCAR_upreg.csv"))
# 
# MF_ovc_cil_new <- read.csv(file = paste0(source, "/cil_GO_MF_OVCAR_upreg.csv"))


#####---CORRECTED GO ANALYSIS---#####


# source of all related .xlsx files
source <- paste0(getwd(), "/corrected_GO_analysis/")

# Read in the .xlsx files to be used.

BP_OVC_sec_new <- read_xlsx(path = paste0(source, "/sec_upreg_OVCAR3_unique.xlsx"), sheet = "GO_BP") |> dplyr::rename(Fold.Enrichment = "Fold Enrichment", Enrichment.FDR = "Enrichment FDR")

CC_PBS_sec_new <- read_xlsx(path = paste0(source, "/sec_downreg_OVCAR3_unique.xlsx"), sheet = "GO_CC") |>  dplyr::rename(Fold.Enrichment = "Fold Enrichment", Enrichment.FDR = "Enrichment FDR")

BP_OVC_cil_new <- read_xlsx(path = paste0(source, "/cil_upreg_OVCAR3_unique.xlsx"), sheet = "GO_BP") |>  dplyr::rename(Fold.Enrichment = "Fold Enrichment", Enrichment.FDR = "Enrichment FDR")

MF_ovc_cil_new <- read_xlsx(path = paste0(source, "/cil_upreg_OVCAR3_unique.xlsx"), sheet = "GO_MF") |>  dplyr::rename(Fold.Enrichment = "Fold Enrichment", Enrichment.FDR = "Enrichment FDR")



```

#### 2.3 Load Analysis Files

```{r}
# 
# cil_v_sec_D1 <- read_xlsx(path = paste0(getwd(), "/analysis/1Day_CILIATED.vs.SECRETORY_segment.comparison_ttest_wilcox.xlsx"), sheet = "All.marker_Cil.vs.Sec") 
# 
# cil_v_sec_D14 <- read_xlsx(path = paste0(getwd(), "/analysis/14Day_CILIATED.vs.SECRETORY_segment.comparison_ttest_wilcox.xlsx"), sheet = "All.marker_Cil.vs.Sec") 
# 
# 
# # Secretory 1 Day
# 
# sec_d1_path <- paste0(getwd(), "/analysis/1Day_Secretory_Trt.comparison_ttest_wilcox.xlsx")
# 
# sec_D1_PBS.vs.FT240 <- read_xlsx(path = sec_d1_path, sheet = "PBS.vs.FT240") 
# sec_D1_FT240.vs.OVCAR3 <- read_xlsx(path = sec_d1_path, sheet = "FT240.vs.OVCAR3") 
# sec_D1_PBS.vs.OVCAR3 <- read_xlsx(path = sec_d1_path, sheet = "PBS.vs.OVCAR3") 
# 
# 
# # Ciliated 1 Day
# 
# cil_d1_path <- paste0(getwd(), "/analysis/1Day_Ciliated_Trt.comparison_ttest_wilcox.xlsx")
# 
# cil_D1_PBS.vs.FT240 <- read_xlsx(path = cil_d1_path, sheet = "PBS.vs.FT240") 
# cil_D1_FT240.vs.OVCAR3 <- read_xlsx(path = cil_d1_path, sheet = "FT240.vs.OVCAR3") 
# cil_D1_PBS.vs.OVCAR3 <- read_xlsx(path = cil_d1_path, sheet = "PBS.vs.OVCAR3") 
# 
# 
# #secretory day 14
# 
# sec_d14_path <- paste0(getwd(), "/analysis/14Day_Secretory_Trt.comparison_ttest_wilcox.xlsx")
# 
# sec_D14_PBS.vs.FT240 <- read_xlsx(path = sec_d14_path, sheet = "PBS.vs.FT240") 
# sec_D14_FT240.vs.OVCAR3 <- read_xlsx(path = sec_d14_path, sheet = "FT240.vs.OVCAR3") 
# sec_D14_PBS.vs.OVCAR3 <- read_xlsx(path = sec_d14_path, sheet = "PBS.vs.OVCAR3") 
# 
# 
# 
# # Ciliated Day 14
# 
# cil_d14_path <- paste0(getwd(), "/analysis/14Day_Ciliated_Trt.comparison_ttest_wilcox.xlsx")
# 
# cil_D14_PBS.vs.FT240 <- read_xlsx(path = cil_d14_path, sheet = "PBS.vs.FT240")
# cil_D14_FT240.vs.OVCAR3 <- read_xlsx(path = cil_d14_path, sheet = "FT240.vs.OVCAR3") 
# cil_D14_PBS.vs.OVCAR3 <- read_xlsx(path = cil_d14_path, sheet = "PBS.vs.OVCAR3") 

```

## Figure 3


This figure contains four volcano plots, all from the 1-day treatment condition:

A - OVCAR3 v PBS in Secretory Cells
B - FT240 v PBS in Secretory Cells
C - OVCAR3 v PBS in Ciliated Cells
D - FT240 v PBS in Ciliated Cells

Note that the Volcano Plots generated may differ slightly from the volcano plots in the paper because tasks such as labeling indivudal points are subject to slight random variations.


#### FUNCT: top_genes()

Usage: this function takes a df of genes with log_fold and p_values and outputs a df with the top n genes according the selected sorting criteria.

Notes about sorting methods:

* sorting by `sig` is straightforward. Top genes are those with the lowest p_value. 
* sorting by `fc` is similarly simple. Top genes are those with the largest |FC|
* the euclidean distance (`euc`) measures the straight line distance between a point and the origin. $D(x,y) = \sqrt{(x_1-x_2)^2 + (y_1-y_2)^2} = \sqrt{x_1^2-y_1^2}$
* Manhattan distance (`manh`) is trickier and measures the distance between two points measured along axes at right angles. (On an xy plane, imagine you can get to a point only by moving in the x or y direction in a straight line. The minimum distance you must travel from point 1 to point 2 in this scenario is the manhattan distance)
** The formula for Manhattan distance is: $D(x,y) = |x_1 -x_2|+|y_1 - y_2| = |x_1| + |y_1| $


Euclidean and Manhattan distances have the advantage of taking into account both significance and fold change when deciding which points are top, which helps highlight the most interesting genes. 


The following code snippet is based on the  methods applied in the R Shiny application VolcaNoseR.


```{r}
top_genes <- function(df = NA, sort_method = "manh", n_genes = 10){
  
  if(sort_method == "manh"){
  # Make a new column called man with the manhattan distance
  df <- df |> mutate(man = abs(log_fold) + abs(minus_log10.p.value_t)) |> arrange(desc(man))}
  else if(sort_method == "euc"){
  # Make a new column called man with the Euclidian distance
  df <- df |> mutate(euc = sqrt(log_fold^2 + minus_log10.p.value_t^2)) |> arrange(desc(euc))
  }
  else if (sort_method == "fc"){
    df <- df |> arrange(desc(abs(log_fold)))
  }
  else if(sort_method == "sig"){
    df <- df |> arrange(desc(minus_log10.p.value_t))
  }
  else{
    message("Not a valid sort method. Select `mnh`, `euc`, `fc`, or `sig`.")
  }
  
  df <- df[abs(df$log_fold) > 0.5 & df$minus_log10.p.value_t > 1.3, ]
  df_out <- head(df, n = n_genes)
  
  return(df_out)
}
```




#### FUNCT: volcano_plot

The following function specifies the volcano plot settings for all the following graphs. 


```{r}


volcano_plot <- function(
    df = NULL,
    highlight_genes_purple = c(), #empty list
    highlight_genes_orange = c(), #empty list
    title_name = NULL,
    upreg = NULL,
    downreg = NULL,
    xmin = -3,
    xmax = 3,
    lab_size = 3,
    FC_cutoff = 0.5,
    flip_log_fold = FALSE,
    num_genes = 20,
    sort_method = "manh"
){
  if(is.null(upreg) & is.null(downreg)){
    FC_name = TeX("$log_2(FC)$")
  }
  else{
    FC_name = TeX(paste0("$log_2(\\frac{\\mu_{", upreg, "}}{\\mu_{", downreg, "}})$"))
  }
  if(is.null(title_name)){
    title_name = deparse(substitute(df))
  }
  if(flip_log_fold){
    df$log_fold <- -df$log_fold
  }
  df$color <- "NS or FC < 0.5"

  df$color[df$log_fold > FC_cutoff & df$p_value_t < 0.05] <- "Upregulated"
  df$color[df$log_fold < -FC_cutoff & df$p_value_t < 0.05] <- "Downregulated"
  # df$color[df$log_fold > FC_cutoff & df$padj < 0.05] <- "Upreg & padj<0.05"
  # df$color[df$log_fold < -FC_cutoff & df$padj < 0.05] <- "Downreg & padj<0.05"
  df$color[df$Marker.name %in% highlight_genes_orange] <- "Orange"
  df$color[df$Marker.name %in% highlight_genes_purple] <- "Purple"
  df$color <- factor(df$color,
                        levels = c("NS or FC < 0.5", "Upregulated", "Downregulated", "Upreg & padj<0.05", "Downreg & padj<0.05", "Orange", "Purple"))

  # df of the top `n` genes to label
  genes_to_label <- top_genes(df = df, sort_method = sort_method, n_gene = num_genes)
  # df of genes to highlight
  genes_to_highlight_orange <- df[df$Marker.name %in% highlight_genes_orange, ]
  genes_to_highlight_purple <- df[df$Marker.name %in% highlight_genes_purple, ]


  ggplot(df, aes(x = log_fold, y = minus_log10.p.value_t))+
    geom_point(aes(color = color)) +
    labs(title = title_name, x = FC_name, y = TeX("$-log_{10}(p_{value})$"))+
    xlim(xmin, xmax)+
    theme(plot.title = element_text(size = 8),
          axis.title = element_text(size = 7))+

    # coord_cartesian(ylim= c(0, 10))+

    scale_color_manual(values = c(`Downregulated` = "blue",
                                  `Upregulated` = "red",
                                  # `Downreg & padj<0.05` = "light blue",
                                  # `Upreg & padj<0.05` = "light red",
                                  `NS or FC < 0.5` = "gray",
                                  `Purple` = "darkorchid",
                                  `Orange` = "darkorange"),
                       guide = guide_legend(override.aes = list(size = 6)))+

    geom_vline(xintercept = 0.5, linetype = "dashed", color = "blue")+
    geom_vline(xintercept = -0.5, linetype = "dashed", color = "blue")+

    # add lines at p-value cut-offs
    geom_hline(yintercept = 1.3, linetype = "dashed", color = "red")+



    # highlight desired genes
        geom_point(data = genes_to_highlight_purple,
               aes(x = log_fold, y = minus_log10.p.value_t),
               color = "darkorchid"
               )+
            geom_point(data = genes_to_highlight_orange,
               aes(x = log_fold, y = minus_log10.p.value_t),
               color = "darkorange"
               )+

    # label upregulated genes

    geom_label_repel(data = genes_to_label,
                     aes(label = Marker.name,
                         x = log_fold,
                         y = minus_log10.p.value_t),
                     # xlim = c(1, NA)
                    # # nudge_x = c(0.5, -0.5),
                    # nudge_y = c(0.2, -0.2),
                    # box.padding = unit(0.5, "lines"),
                    # point.padding = unit(.3+3*0.1, "lines"),
                    # max.time = 100,
                    # max.iter = 10^6,
                    size = lab_size,
                    max.overlaps = 40
                    )+

    geom_label_repel(data = genes_to_highlight_orange,
                     aes(label = Marker.name,
                         x = log_fold,
                         y = minus_log10.p.value_t),
                     # xlim = c(1, NA)
                    # # nudge_x = c(0.5, -0.5),
                    # nudge_y = c(0.2, -0.2),
                    # box.padding = unit(0.5, "lines"),
                    # point.padding = unit(.3+3*0.1, "lines"),
                    # max.time = 100,
                    # max.iter = 10^6,
                    size = lab_size,
                    max.overlaps = 40,
                    colour = "black"
                    )+

    geom_label_repel(data = genes_to_highlight_purple,
                     aes(label = Marker.name,
                         x = log_fold,
                         y = minus_log10.p.value_t),
                     # xlim = c(1, NA)
                    # # nudge_x = c(0.5, -0.5),
                    # nudge_y = c(0.2, -0.2),
                    # box.padding = unit(0.5, "lines"),
                    # point.padding = unit(.3+3*0.1, "lines"),
                    # max.time = 100,
                    # max.iter = 10^6,
                    size = lab_size,
                    max.overlaps = 40,
                    colour = "black"
                    )+

    theme_light(base_size = 16, base_family = "sans")+

    # remove dumb grid-lines
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())+
    labs(color = "Color")+
    theme(legend.position = "none")
}


```



#### FUNCT: add_labels

```{r}

add_labels <- function(plot = NA, df = NA, n_genes = 10, method = "manh", highlight_genes = c(), lab_size = 3, flip_log_fold = F, x_lim_max = 3){
    if(flip_log_fold){
      df$log_fold <- -df$log_fold
      }
    
  
    # find the top n genes using the selected method
    # return names
    top_n <- top_genes(df = df, sort_method = method, n_genes = n_genes)$Marker.name
    
    # merge the name list, removing duplicate values
    gene_list <- unique(c(top_n, highlight_genes))
    
    # filter dataset to get only genes to label
  
    genes_to_label <- filter(df, df$Marker.name %in% gene_list)
  
    plot_out <- plot +

    # Now add labels to both the up and down regulated genes
      
    # for log_fold > 0, plot between 05 and 3

    geom_label_repel(data = subset(genes_to_label, log_fold > 0),
                     aes(label = Marker.name,
                         x = log_fold,
                         y = minus_log10.p.value_t),
                     xlim = c(0.5, x_lim_max),
                     # xlim = c(1, NA)
                    # # nudge_x = c(0.5, -0.5),
                    # nudge_y = c(0.2, -0.2),
                    # box.padding = unit(0.5, "lines"),
                    # point.padding = unit(.3+3*0.1, "lines"),
                    # max.time = 100,
                    # max.iter = 10^6,
                    size = lab_size,
                    max.overlaps = 40
                    )+
      
      # for log_fold < 0, plot -0.5 to -3
      geom_label_repel(data = subset(genes_to_label, log_fold < 0),
                     aes(label = Marker.name,
                         x = log_fold,
                         y = minus_log10.p.value_t),
                     xlim = c(-x_lim_max, -0.5),
                     # xlim = c(1, NA)
                    # nudge_x = c(0.5, -0.5),
                    # nudge_y = c(0.2, -0.2),
                    # box.padding = unit(0.5, "lines"),
                    # point.padding = unit(.3+3*0.1, "lines"),
                    # max.time = 100,
                    # max.iter = 10^6,
                    size = lab_size,
                    max.overlaps = 20
                    )

    
    return(plot_out)
  
    }


```


#### FUNCT: venn_compare()

```{r}

# create a function to produce two Venn Diagram plots -- one comparing upregulated genes and the over comparing downregulated genes.

venn_compare <- function(data_left = sec_D1_PBS.vs.FT240, data_right = sec_D1_PBS.vs.OVCAR3, flip_log_fold = FALSE, left = "FT240", right = "OVCAR3"){

  if(flip_log_fold){
    # flip both log fold signs
    data_left$log_fold <- -data_left$log_fold
    data_right$log_fold <- -data_right$log_fold
  }
  
  
  if(flip_log_fold == "data_left"){
    # flip only data 1
    data_left$log_fold <- -data_left$log_fold
  }
  
  if(flip_log_fold == "data_right"){
    # flip only data 2
    data_right$log_fold <- -data_right$log_fold 
  }
    
  # get a list of all upregulated genes in data1
  upreg_l <- data_left[data_left$log_fold > 0.5 & data_left$p_value_t < 0.05, ]$Marker.name
  # get a list of DOWNregulated in data1
  downreg_l <- data_left[data_left$log_fold < -0.5 & data_left$p_value_t < 0.05, ]$Marker.name
  
  #UP data 2
  upreg_r <- data_left[data_right$log_fold > 0.5 & data_right$p_value_t < 0.05, ]$Marker.name
  #DOWN data 2
  downreg_r <- data_right[data_right$log_fold < -0.5 & data_right$p_value_t < 0.05, ]$Marker.name
  
  
  # create a list of the upregulated genes
  upreg <- setNames(list(upreg_l, upreg_r), c(left, right))
  # create list downreg genes
  downreg <- setNames(list(downreg_l, downreg_r), c(left, right))
    
  # use euler to create the plots  
  euler_up <- euler(upreg)
  euler_down <- euler(downreg)
    
  # Plot the Venn diagram with proportional scaling
  plot_up <- plot(euler_up, 
                  fills = c("darkviolet", "darkgreen"), 
                  alpha = 0.5, 
                  quantities = list(cex = 3), 
                  labels = FALSE) |>
    as_ggplot() # convert to ggplot
  
  
  plot_down <- plot(euler_down, 
                    fills = c("violet", "lightgreen"), 
                    alpha = 0.5, 
                    quantities = list(cex = 3), 
                    labels = FALSE) |>
    as_ggplot()
  

  plot_up <- plot_up +
    theme(plot.margin = unit(c(1, 1, 1, 1), "cm")) +  # Increase margins (top, right, bottom, left)
    annotate("text", x = 0.08, y = 1, label = paste0(left, " \n UP"), size = 8, color = "darkviolet")+
    annotate("text", x = 0.95, y = 1, label = paste0(right, " \n UP"), size = 8, color = "darkgreen")

  plot_down <- plot_down +
    theme(plot.margin = unit(c(1, 1, 1, 1), "cm")) +  # Increase margins (top, right, bottom, left)
    annotate("text", x = 0.08, y = 1, label = paste0(left, " \n DOWN"), size = 8, color = "darkviolet")+
    annotate("text", x = 0.95, y = 1, label = paste0(right, " \n DOWN"), size = 8, color = "darkgreen")

  
  
  combined_plot <- plot_up + plot_down
  
  combined_plot
  
}


```

#### Fig3.1 -  Volcano Plots

```{r fig.height=9, fig.width=13}


# Day 1 Volcano Plots


# choose label size
size <- 3
genes <- 40


# Checks

highlight_up_FT240_sec <-  sec_D1_PBS.vs.FT240[sec_D1_PBS.vs.FT240$log_fold > 0.5 & sec_D1_PBS.vs.FT240$p_value_t < 0.05, ]$Marker.name
highlight_down_FT240_sec <-  sec_D1_PBS.vs.FT240[sec_D1_PBS.vs.FT240$log_fold < -0.5 & sec_D1_PBS.vs.FT240$p_value_t < 0.05, ]$Marker.name
highlight_up_OVCAR3_sec <-  sec_D1_PBS.vs.OVCAR3[sec_D1_PBS.vs.OVCAR3$log_fold > 0.5 & sec_D1_PBS.vs.OVCAR3$p_value_t < 0.05, ]$Marker.name
 <-  sec_D1_PBS.vs.OVCAR3[sec_D1_PBS.vs.OVCAR3$log_fold < -0.5 & sec_D1_PBS.vs.OVCAR3$p_value_t < 0.05, ]$Marker.name

highlight_up_OVCAR3_sec <-  sec_D1_PBS.vs.OVCAR3[sec_D1_PBS.vs.OVCAR3$log_fold > 0.5 & sec_D1_PBS.vs.OVCAR3$p_value_t < 0.05, ]$Marker.name
highlight_down_OVCAR3_sec <-  sec_D1_PBS.vs.OVCAR3[sec_D1_PBS.vs.OVCAR3$log_fold < -0.5 & sec_D1_PBS.vs.OVCAR3$p_value_t < 0.05, ]$Marker.name



# list_changed_OVCAR3_sec <- c("ACTB", "LDHB", "CD44", "NUPR1", "IRF1", "S100A9", "LTF", "KIR Family")
# list_changed_FT240_sec <- c("ALDOA", "CSF3R", "CALML3", "C4B")

# list_genes <- c("CXCL2", "CCL2", "CXCL10", "CCL18", "VCAM1", "FLNA", "CSF1", "ICAM1", "TXNIP")

d1_sec_plot_OVC <- volcano_plot(df = sec_D1_PBS.vs.OVCAR3 , title = "OVCAR3 vs PBS \n(Secretory Cells, Day 1)", 
                                # highlight_genes_purple= c("REL", "TNFAIP3"), #highlight_down_OVCAR3_sec,
                                #highlight_genes_orange=highlight_up_OVCAR3_sec,
                                upreg = "OVCAR3", downreg = "PBS", flip_log_fold = F, lab_size = size, num_genes = genes)+ theme(
  axis.title = element_text(size = 20)) 


d1_cil_plot_OVC <- volcano_plot(df = cil_D1_PBS.vs.OVCAR3 , title = "OVCAR3 vs PBS \n(Ciliated Cells, Day 1)", 
                                highlight_genes_purple= c("REL", "TNFAIP3"),# highlight_genes_purple=c("PPL", "HSP90B1", "FGF10"),
                                # highlight_genes_orange=c("CDKN1C"),
                                upreg = "OVCAR3", downreg = "PBS", flip_log_fold = F, lab_size = size, num_genes = genes) + theme(
  axis.title = element_text(size = 20))





d1_sec_plot_FT <- volcano_plot(df = sec_D1_PBS.vs.FT240, title = "FT240 vs PBS \n(Secretory Cells, Day 1)", 
                                # highlight_genes_purple= c("ALDOA", "CSF3R", "CALML3", "C4B"), 
                                # highlight_genes_orange= c("CCL3/L1"),
                               upreg = "FT240", downreg = "PBS", flip_log_fold = F, lab_size = size, num_genes = genes)+ theme(
  axis.title = element_text(size = 20)) 


d1_cil_plot_FT <- volcano_plot(df = cil_D1_PBS.vs.FT240 , title = "FT240 vs PBS \n(Ciliated Cells, Day 1)", 
                                # highlight_genes_purple=c("ACTB", "CD63", "CALML3"),
                                # highlight_genes_orange=c("CD68", "PTTG2", "CCL15"),
                               upreg = "FT240", downreg = "PBS", flip_log_fold = F, lab_size = size, num_genes = genes) + theme(
  axis.title = element_text(size = 20))



  (d1_sec_plot_OVC | d1_sec_plot_FT) / 
  (d1_cil_plot_OVC | d1_cil_plot_FT) 

```

#### Fig3.2 - Venn Diagrams


```{r fig.height=3.5, fig.width=10}


# Day 1 Venn Diagrams

venn_d1_sec <- venn_compare(data_left = sec_D1_PBS.vs.FT240, data_right = sec_D1_PBS.vs.OVCAR3, 
                            flip_log_fold = F, left = "FT240", right = "OVCAR3")


venn_d1_cil <- venn_compare(data_left = cil_D1_PBS.vs.FT240, data_right = cil_D1_PBS.vs.OVCAR3, 
                            flip_log_fold = F, left = "FT240", right = "OVCAR3") 


venn_d1_sec

venn_d1_cil

```


## Figure 4


This figure contains four volcano plots, all from the 14-day treatment condition:

A - OVCAR3 v PBS in Secretory Cells
B - FT240 v PBS in Secretory Cells
C - OVCAR3 v PBS in Ciliated Cells
D - FT240 v PBS in Ciliated Cells

Note that the Volcano Plots generated may differ slightly from the volcano plots in the paper because tasks such as labeling indivudal points are subject to slight random variations.

#### Fig4.1 - Volcano plots

```{r fig.height=9, fig.width=13}


# Day14 Volcano plots


d14_sec_plot_OVC <- volcano_plot(df = sec_D14_PBS.vs.OVCAR3 , title = "OVCAR3 vs PBS \n(Secretory Cells, Day 14)", upreg = "OVCAR3", downreg = "PBS", flip_log_fold = F, lab_size = size)+ theme(
  axis.title = element_text(size = 20))


d14_cil_plot_OVC <- volcano_plot(df = cil_D14_PBS.vs.OVCAR3 , title = "OVCAR3 vs PBS \n(Ciliated Cells, Day 14)", upreg = "OVCAR3", downreg = "PBS", flip_log_fold = F, lab_size = size) + theme(
  axis.title = element_text(size = 20))


d14_sec_plot_FT <- volcano_plot(df = sec_D14_PBS.vs.FT240, title = "FT240 vs PBS \n(Secretory Cells, Day 14)", upreg = "FT240", downreg = "PBS", flip_log_fold = F, lab_size = size)+ theme(
  axis.title = element_text(size = 20))


d14_cil_plot_FT <- volcano_plot(df = cil_D14_PBS.vs.FT240 , title = "FT240 vs PBS \n(Ciliated Cells, Day 14)", upreg = "FT240", downreg = "PBS", flip_log_fold = F, lab_size = size) + theme(
  axis.title = element_text(size = 20))




  (d14_sec_plot_OVC | d14_sec_plot_FT )  /
  (d14_cil_plot_OVC | d14_cil_plot_FT)


```

#### Fig4.1 - Venn Diagrams

```{r fig.height=3.5, fig.width=10}

# Day 14 Venn Diagrams


venn_d14_sec <- venn_compare(data_left = sec_D14_PBS.vs.FT240, data_right = sec_D14_PBS.vs.OVCAR3, 
                            flip_log_fold = F, left = "FT240", right = "OVCAR3") 

venn_d14_cil <- venn_compare(data_left = cil_D14_PBS.vs.FT240, data_right = cil_D14_PBS.vs.OVCAR3, 
                            flip_log_fold = F, left = "FT240", right = "OVCAR3") 


venn_d14_sec

venn_d14_cil

```

## Figure 5

Figure 5 included 4 GO plots and 2 heat maps showing upregulated and downregulated genes following OVCAR3 EV treatment. 


### Fig5.1 - Create GO Plots

#### FUNCT: plot_go()

The following function creates plots to display the results of Gene Ontology analysis data.

It takes a table with Pathway, Fold.Enrichment,  nGenes, and Enrichment.FDR and outputs the GO plot. 

```{r}

plot_GO <- function(data = BP_OVC_sec, n_graph = 10, n_genes_min = 5){
  
  # retrieve the top n pathways
  data <- data[data$nGenes > n_genes_min, ]
    
  plot1_data <- head(data, n = n_graph) 
  

  
  # simplify pathway names
  plot1_data$Pathway <- gsub("^GO:[0-9]+ ", "", plot1_data$Pathway)
  
  # Define a list of abbreviations
  abbreviations <- c("response" = "resp", 
                     "mediated by" = "med", 
                     "chemotaxis" = "chemotx", 
                     "migration" = "migr", 
                     "immune" = "imm", 
                     "antimicrobial" = "antimicro",
                     "differentiation" = "dif",
                     "regeneration" = "regen",
                     "proliferation" = "prolif",
                     "negative" = "neg",
                     "positive" = "pos",
                     "regulation" = "reg",
                     "binding" = "bind.",
                     "transcription" = "transcr",
                     "regulatory" = "reg")

  # Function to replace words with abbreviations
  simplify_names <- function(name) {
      for (word in names(abbreviations)) {
        name <- gsub(word, abbreviations[[word]], name)
      }
      return(name)
    }

  # Apply the function to the Pathway column to simplify names
  plot1_data$Pathway <- sapply(plot1_data$Pathway, simplify_names)
    


  # sort by Fold.Enrichment
  plot_df <- plot1_data %>%
  arrange(Fold.Enrichment) |> mutate(Pathway = factor(Pathway, levels = Pathway))
  
  # Use the WesAnderson package to create a continous palette
  pal <- wes_palette("Zissou1", 100, type = "continuous")


  # create the initial plot
  init_plot <- ggplot(data = plot_df,  aes(x=Fold.Enrichment, y= Pathway)) +
        geom_segment(aes(x= 0, xend=Fold.Enrichment), linewidth = 1) +
        geom_point(aes(size= nGenes, color= Enrichment.FDR)) +
        theme_bw() +
        scale_color_gradientn(colors = rev(pal))+
        scale_y_discrete(labels = function(x) str_wrap(x, width = 25))+
      labs(x = "Fold Enrichment")
    
  
  # Adjust title size and element sizes
    go_plot <- init_plot + theme(
      axis.title=element_text(size=24),
      axis.text=element_text(size=18),
      plot.title=element_text(size=18),
      legend.text =element_text(size=18),
      legend.title = element_text(size = 18),
      axis.text.y = element_text(vjust = 0.5))
    
    return(go_plot)
}
```


####  create_GO_plots

These GO plots use the new GO tables generated using just the OVCAR3 upregulated genes that were NOT also upregulated by FT240 treatment (i.e. uniquely upregulated transcripts).


```{r}


BP_SEC_GO <- plot_GO(data = BP_OVC_sec_new, n_graph = 15) + ggtitle("GO Biological Pathways upregulated in \n Secretory Cells after 1-Day Treatment \n with OVCAR3 EVs") + theme(plot.caption = element_text(hjust = 0, face= "italic"), #Default is hjust=1
        plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot")+ theme(plot.title = element_text(size = 24))

CC_SEC_GO <- plot_GO(data = CC_PBS_sec_new, n_graph = 15)  + ggtitle("GO Cellular Components downregulated in \n Secretory Cells after 1-Day Treatment \n with OVCAR3 EVs
") + theme(plot.caption = element_text(hjust = 0, face= "italic"), #Default is hjust=1
        plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot")+ theme(plot.title = element_text(size = 24))


BP_CIL_GO <- plot_GO(data = BP_OVC_cil_new, n_graph = 15) + ggtitle("GO Biological Pathways upregulated in \n Ciliated Cells after 1-Day Treatment \n with OVCAR3 EVs") + theme(plot.caption = element_text(hjust = 0, face= "italic"), #Default is hjust=1
        plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot")+ theme(plot.title = element_text(size = 24))


MF_CIL_GO <- plot_GO(data = MF_ovc_cil_new, n_graph = 10) + ggtitle("GO Molecular Function upregulated in \n Ciliated Cells after 1-Day Treatment \n with OVCAR3 EVs") + theme(plot.caption = element_text(hjust = 0, face= "italic"), #Default is hjust=1
        plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot")+ theme(plot.title = element_text(size = 24))



```

####  Merge GO plots into one figure

```{r fig.height=24, fig.width=16}

# Generate the GO Plots for the figure

figure_right <- (BP_SEC_GO  + CC_SEC_GO + BP_CIL_GO + MF_CIL_GO)

figure_right


```


### Fig5.2 - Create heatmap

Create heatmaps with

1. Patients in X-axis
2. Genes of interest in y-axis
3. values of genes of interest filled in as colors. 

#### FUNCT: subset_dataset
```{r}
subset_dataset <- function(dataset = NA, subset_by = NA, value_to_select = NA){
  
    # check to see if subsetting requested
  if(is.character(subset_by)){
    # check to see if a value provided. 
    if(!is.character(value_to_select)){
      message("No value provided to subset on. `value_to_select` must be a character.")
    }
    else{
      # in the column subset_by
      # select all of the rows that match "value_to_select"
      dataset = dataset[, pData(dataset)[, subset_by] %in% value_to_select] 
    }
  }
  
  return(dataset)
}
```




```{r}


# create a dataset with only 1-Day data

EV_1_day <- subset_dataset(dataset = EV_q_norm, subset_by = "Days", value_to_select = c("1-Day"))

# create a dataset with only Secretory Cell data

EV_1_day_sec <- subset_dataset(dataset = EV_1_day, subset_by = "segment", value_to_select = c("Secretory"))

# 1 day, secretory, remove FT240 treatment condition 

EV_1_day_sec_noFT240 <- subset_dataset(dataset = EV_1_day_sec, subset_by = "Treatment", value_to_select = c("OVCAR3", "PBS"))


# create a subset with only secretory cell data

EV_1_day_cil <- subset_dataset(dataset = EV_1_day, subset_by = "segment", value_to_select = c("Ciliated"))
# 1 day, secretory, remove FT240 treatment condition 

EV_1_day_cil_noFT240 <- subset_dataset(dataset = EV_1_day_cil, subset_by = "Treatment", value_to_select = c("OVCAR3", "PBS"))

```




```{r}
# 
# 
# # Secretory 1 Day
# # 
# sec_d1_path <- paste0(getwd(), "/analysis/1Day_Secretory_Trt.comparison_ttest_wilcox.xlsx")
# 
# sec_D1_PBS.vs.FT240 <- read_xlsx(path = sec_d1_path, sheet = "PBS.vs.FT240")
# # sec_D1_FT240.vs.OVCAR3 <- read_xlsx(path = sec_d1_path, sheet = "FT240.vs.OVCAR3")
# sec_D1_PBS.vs.OVCAR3 <- read_xlsx(path = sec_d1_path, sheet = "PBS.vs.OVCAR3")
# 
# 
# # # Ciliated 1 Day
# 
# cil_d1_path <- paste0(getwd(), "/analysis/1Day_Ciliated_Trt.comparison_ttest_wilcox.xlsx")
# 
# cil_D1_PBS.vs.FT240 <- read_xlsx(path = cil_d1_path, sheet = "PBS.vs.FT240")
# # cil_D1_FT240.vs.OVCAR3 <- read_xlsx(path = cil_d1_path, sheet = "FT240.vs.OVCAR3")
# cil_D1_PBS.vs.OVCAR3 <- read_xlsx(path = cil_d1_path, sheet = "PBS.vs.OVCAR3")


```


```{r}

# #secretory day 14
# 
# sec_d14_path <- paste0(getwd(), "/analysis/14Day_Secretory_Trt.comparison_ttest_wilcox.xlsx")
# 
# # sec_D14_PBS.vs.FT240 <- read_xlsx(path = sec_d14_path, sheet = "PBS.vs.FT240") 
# # sec_D14_FT240.vs.OVCAR3 <- read_xlsx(path = sec_d14_path, sheet = "FT240.vs.OVCAR3") 
# sec_D14_PBS.vs.OVCAR3 <- read_xlsx(path = sec_d14_path, sheet = "PBS.vs.OVCAR3") 
# 
# 
# 
# # Ciliated Day 14
# 
# cil_d14_path <- paste0(getwd(), "/analysis/14Day_Ciliated_Trt.comparison_ttest_wilcox.xlsx")
# 
# # cil_D14_PBS.vs.FT240 <- read_xlsx(path = cil_d14_path, sheet = "PBS.vs.FT240")
# # cil_D14_FT240.vs.OVCAR3 <- read_xlsx(path = cil_d14_path, sheet = "FT240.vs.OVCAR3") 
# cil_D14_PBS.vs.OVCAR3 <- read_xlsx(path = cil_d14_path, sheet = "PBS.vs.OVCAR3")


```


####  List Genes of Interest (GOI)

```{r}

pt_IDs <- unique(pData(EV_1_day_cil)$Patient)

treatments <- unique(pData(EV_1_day_cil)$Treatment)

GOI_up <- sec_D1_PBS.vs.OVCAR3[(sec_D1_PBS.vs.OVCAR3$log_fold > 0.5 & sec_D1_PBS.vs.OVCAR3$p_value_t < 0.05), ]$Marker.name

GOI_down <- sec_D1_PBS.vs.OVCAR3[(sec_D1_PBS.vs.OVCAR3$log_fold < -0.5 & sec_D1_PBS.vs.OVCAR3$p_value_t < 0.05), ]$Marker.name

GOI <-  c(GOI_up, GOI_down)


# function to get significant genes
get_sig_genes <- function(df, type = "upreg", flip_log_fold = F,  lfc_thresh = 0.5, p_thresh = 0.05){
  
  # flip log fold change if needed
  if(flip_log_fold){
    df$log_fold = -df$log_fold
  }
  
  if(type == "upreg"){
    sig_genes <- df[df$log_fold > lfc_thresh & df$p_value_t < p_thresh, ]$Marker.name
  }
  else if(type == "downreg"){
    sig_genes <- df[df$log_fold < -lfc_thresh & df$p_value_t < p_thresh, ]$Marker.name
  }
  
  return(sig_genes)
  
}

# now create list of just OVCAR3 Up and DOWN

# Secretory Segments

  upreg_OVC_sec <- get_sig_genes(df = sec_D1_PBS.vs.OVCAR3, type = "upreg", flip_log_fold = F)
  
  downreg_OVC_sec <- get_sig_genes(df = sec_D1_PBS.vs.OVCAR3, type = "downreg", flip_log_fold = F)

  upreg_FT240_sec <- get_sig_genes(df = sec_D1_PBS.vs.FT240, type = "upreg", flip_log_fold = F)
    
  downreg_FT240_sec  <- get_sig_genes(df = sec_D1_PBS.vs.FT240, type = "downreg", flip_log_fold = F)
  
  # Ciliated segments
  
  upreg_OVC_cil <- get_sig_genes(df = cil_D1_PBS.vs.OVCAR3, type = "upreg", flip_log_fold = F)
  
  downreg_OVC_cil <- get_sig_genes(df = cil_D1_PBS.vs.OVCAR3, type = "downreg", flip_log_fold = F)

  upreg_FT240_cil <- get_sig_genes(df = cil_D1_PBS.vs.FT240, type = "upreg", flip_log_fold = F)
    
  downreg_FT240_cil  <- get_sig_genes(df = cil_D1_PBS.vs.FT240, type = "downreg", flip_log_fold = F)
  


  
  # Genes of interest, Secretory
upreg_OVC_only_sec <- upreg_OVC_sec[!(upreg_OVC_sec %in% upreg_FT240_sec)] # length = 36
  
downreg_OVC_only_sec <- downreg_OVC_sec[!(downreg_OVC_sec %in% downreg_FT240_sec)] # length = 25
  
  
GOI_sec <- c(upreg_OVC_only_sec, downreg_OVC_only_sec) # 36, 22 


  

  upreg_OVC_only_cil <- upreg_OVC_cil[!(upreg_OVC_cil %in% upreg_FT240_cil)] # length = 13
  
  downreg_OVC_only_cil <- downreg_OVC_cil[!(downreg_OVC_cil %in% downreg_FT240_cil)] # length = 8
  
  
GOI_cil <- c(upreg_OVC_only_cil, downreg_OVC_only_cil) # 13, 6

# agrees with venn diagram

```


####  Functions for Heatmap

```{r}

pt_order <- c("pt9", "pt12", "pt10", "pt6", "pt1", "pt3", "pt5", "pt7", "pt2", "pt11", "pt8", "pt4", "pt13")
  #c("UH000025", "5929-606856", "5929-606939", "NW71422", "461-27304", "5929-607267", "5929-607140", "461-27413", "5929-606950", "461-27444", "UH00006", "5929-606994")



create_annotation_col <- function(dataset = EV_1_day_sec, 
                                  order_treatment = c("PBS", "OVCAR3"), 
                                  order_Pts = pt_order){
  
  # create the annotation column with only Treatment and Patient info

  annotation_col = pData(dataset)[, c("Treatment", "Patient"),]
  
  # factor the Treatment and Patient columns
  # arrange based on order provided
  
  annotation_col <- annotation_col |> mutate(
  Treatment = factor(Treatment, levels = order_treatment), # factor Treatment
  Patient = factor(Patient, levels = order_Pts)) |> # factor Patient
    dplyr::arrange(Treatment, Patient) #arrange based on factors


  return(annotation_col)
  
}



create_ordered_matrix <- function(dataset = EV_1_day_sec, 
                                  order_treatment = c("PBS", "FT240", "OVCAR3"), 
                                  order_Pts = pt_order, 
                                  gene_list = GOI_new){
  
  # create matrix of all gene data based on genes of interest

  matrix <- assayDataElement(dataset[gene_list, ], elt = "q_norm")
  
  
  annotation_col <- create_annotation_col(dataset = dataset, order_treatment = order_treatment, order_Pts = order_Pts)
  
  # determine the order of the columns in the matrix

  order_cols <- rownames(annotation_col)
  
  
  # Reorder the columns of the matrix according to 'order'
  matrix_reordered <- matrix[, order_cols]
  
  return(matrix_reordered)
  
}


```


####  Create Colors for heatmap


```{r fig.height=12, fig.width=6}

# Heatmap colors are stored as
# 
# colors$Name_Column <- c("Color1", "Color2") 
# 
# names(colors$Name_Column) <- c("Item1", "Item2")


#### CREATE Colors  ####

# select colors for Treatments
my_colors <- c("forestgreen", "darkviolet", "navyblue")
# assign Treatment names
names(my_colors) <- c("OVCAR3", "FT240", "PBS")


# store as a list with $Treatment
my_colors <- list(Treatment = my_colors)
# alternative rainbow color scheme
# new_cols <- colorRampPalette(grDevices::rainbow(length(unique(pData(EV_1_day_sec_noFT240)$Tissue))))

# use discrete rainbow from `kroma` to produce rainbow colors
my_colors$Patient <- rev(khroma::color("discrete rainbow")(length(unique(pData(EV_1_day_sec)$Patient))))

# assign names used the pt_order vectro

names(my_colors$Patient) <- pt_order


#### CREATE Colors  ####


```


#### FUNCT: create_heatmap()

This function used `ComplexHeatmap` to create a heatmap. 

```{r}

create_heatmap <- function(dataset = EV_1_day_sec, 
                                  order_treatment = c("PBS", "FT240", "OVCAR3"), 
                                  order_Pts = pt_order,
                                  annotation_colors = my_colors, 
                                  title = "Insert Title Here",
                                  legend = TRUE, 
                                  font = 20,
                                  gene_list = GOI_new,
                                  split_point = 36,
                                  num_PBS = 3,
                                  num_FT240 = 2){
  
  # create the annotation column
  annotation_col <- create_annotation_col(dataset = dataset, order_treatment = order_treatment, order_Pts = order_Pts)
  # create the matrix with expression data
  matrix_reordered <- create_ordered_matrix(dataset = dataset, order_treatment = order_treatment, order_Pts = order_Pts, gene_list = gene_list)
  
  
  # First, convert your matrix to a data frame for easier manipulation
  mat_df <- as.data.frame(matrix_reordered)

  # Transpose the matrix so that columns become rows (for easier merging with annotation)
  mat_df <- t(mat_df)
  
  

  # Merge the matrix data with the annotation data based on column names
  merged_df <- merge(mat_df, annotation_col, by = "row.names" )
  
  # Now, group by treatment and patient and compute the row-wise average
  averaged_df <- merged_df %>%
    group_by(Treatment, Patient) %>%
    summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))  # Average each row, excluding non-numeric columns
  
  
  # create new annotation df
  
  annotation_df <- as.data.frame(averaged_df[1:2])
  
  matrix_averaged <- t(as.matrix(averaged_df[3: ncol(averaged_df)]))
  
  colnames(matrix_averaged) <- rownames(annotation_df)
  
  
  # implement row scaling - z-score normalization across row
  matrix_scaled <- t(scale(t(matrix_averaged)))
  
  
  # create complext heatmap object 
  
  heatmap = Heatmap(
    matrix_scaled,
    name = title,
    col = colorRamp2(
    c(-3.5, 0, 3.5),  # Set color at 0 to white
    c("navy", "white", "firebrick3")),
  
  # colorRampPalette(c("navy", "white", "firebrick3"))(21),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_split = factor(
                c(rep("UP OVCAR3", split_point), rep("DOWN OVCAR3", nrow(matrix_scaled) - split_point)),
                levels = c("UP OVCAR3", "DOWN OVCAR3")  # Specify the desired order
                ),  # Adjust row gaps
  
    column_split = factor(
                  c(rep("PBS", num_PBS), rep("FT240", num_FT240), rep("OVCAR3", ncol(matrix_scaled) - num_PBS - num_FT240)),
                    levels = c("PBS", "FT240", "OVCAR3")),
  
    top_annotation = HeatmapAnnotation(df = annotation_df, col = annotation_colors, 
                                       annotation_name_gp = gpar(fontsize = font),
                                       show_annotation_name = legend, # turn off annotation name if legend = F
                                       annotation_legend_param = list(
                                         ncol = 4, # Set the number of columns for the legend
                                         gp = gpar(fontsize = font),
                                         title_gp = gpar(fontsize = font + 2), # title font size
                                         labels_gp = gpar(fontsize = font)   # Label font size  
      )),  # Top annotations
    column_title = title, # add a title to the heatmap
    show_heatmap_legend = legend,
    show_row_names = TRUE,
    show_column_names = FALSE,
    border = TRUE,
    row_gap = unit(1, "mm"),  # Customize row gaps
    heatmap_legend_param = list(
      title = "Z-score", legend_direction = "horizontal", side = "right", title_gp = gpar(fontsize = font+2),  # Title font size
    labels_gp = gpar(fontsize = font)   # Label font size
    ),
  
    row_names_gp = gpar(fontsize = font-8),
    row_title_gp = gpar(fontsize = font+8),   # Row title font size
    column_title_gp = gpar(fontsize = font+8) # Column title font size
  )


  
  return(heatmap)
}



```



#### Create Heatmaps and Combine

```{r fig.height=20, fig.width=12}


heatmap1 = create_heatmap(dataset = EV_1_day_sec, 
                  order_treatment = c("PBS", "FT240", "OVCAR3"), 
                  order_Pts = pt_order, 
                  annotation_colors = my_colors,
                  title = "1-Day Treatment, \n Secretory Cells", legend = FALSE, font = 24,
                  gene_list = GOI_sec,
                  split_point = 50,
                  num_PBS = 12,
                  num_FT240 = 11)


heatmap2 = create_heatmap(dataset = EV_1_day_cil, 
                      order_treatment = c("PBS", "FT240", "OVCAR3"), 
                      order_Pts = pt_order, 
                      annotation_colors = my_colors,
                      title = "1-Day Treatment, \n Ciliated Cells", font = 24,
                      gene_list = GOI_cil,
                      split_point = 21,
                      num_PBS = 11,
                      num_FT240 = 10)






gb = grid.grabExpr(draw(
  heatmap1, 
  heatmap_legend_side = "bottom", 
  annotation_legend_side = "bottom", 
  merge_legend = TRUE,
))

gb2 = grid.grabExpr(draw(
  heatmap2, 
  heatmap_legend_side = "bottom", 
  annotation_legend_side = "bottom", 
  merge_legend = TRUE,
))

library(gridExtra)

grid.arrange(gb, gb2, ncol = 2)

# draw(
#   combined_heatmap,
#   heatmap_legend_side = "bottom", 
#   annotation_legend_side = "bottom", 
#   merge_legend = TRUE
#   )
# # 
# draw(
#   heatmap2, 
#   heatmap_legend_side = "bottom", 
#   annotation_legend_side = "bottom", 
#   merge_legend = TRUE
#   )
# 



```


### Final Figure 5

```{r fig.height=20, fig.width=30}

(figure_right | wrap_elements(gb) | wrap_elements(gb2)) + plot_layout(widths = c(12, 5, 5))

```


## Figure 6

Figure 6 analyzes data from Mass spectrophotometry analysis of EV protein. 

The files needed to generate these plots are found under ~analysis/proteomics

Figure6B - Volcano plot comparing EV protein change in OVCAR3 vs PBS treated samples

Figure6C - Volcano plot comparing EV protein change in OVCAR3 vs FT240 treated Samples

Figure6F - GO biological Pathways associated with exo-proteins induced by OVCAR3 EVs

### Fig6.1 - Volcano Plots


#### Load Files VolcanoPlot

```{r}


my_comparisons = list(
  c("FT.C", "p53.C"),
  c("p53.C", "STIC.C"),
  c("STIC.C", "Cancer"), 
  c("p53.C", "Cancer"),
  c("FT.C", "Cancer")
)


add_sig <- function(plot, comparisons_wanted = my_comparisons, pnt = 5){
  plot <- plot + 
    stat_compare_means(
      comparisons = comparisons_wanted, 
      method = "t.test", 
      bracket.nudge.y = -4,
      size = pnt) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.10)))
  return (plot)
}

```



```{r}

source_path <- paste0(getwd(), "/analysis/proteomics")



EV_protein_OVCAR3 <- read_xlsx(path = paste0(source_path, "/PBS.vs.OVCAR3_comparison.xlsx"))


EV_protein_FT240 <- read_xlsx(path = paste0(source_path, "/PBS.vs.FT240_comparison.xlsx"))


EV_protein_OVCAR3 <- EV_protein_OVCAR3 |> dplyr::rename(c(p_value_t = p_value_ttest, minus_log10.p.value_t = minuslog10.p.value.ttest))

EV_protein_FT240 <- EV_protein_FT240 |> dplyr::rename(c(p_value_t = p_value_ttest, minus_log10.p.value_t = minuslog10.p.value.ttest))

```


#### Generate volcano plot


```{r fig.height=8, fig.width=12}

VP_OVCAR3 <- volcano_plot(df = EV_protein_OVCAR3, xmin = -6, xmax = 6, num_genes = 0, title_name = "hFTE EV proteome change following EV insult", upreg = "OVCAR3", downreg = "PBS", flip_log_fold = T) |> add_labels(df = EV_protein_OVCAR3, n_genes = 40, lab_size = 6, x_lim_max = 6, flip_log_fold = T)

VP_FT240 <- volcano_plot(df = EV_protein_FT240, xmin = -6, xmax = 6, num_genes = 0, title_name = "hFTE EV proteome change following EV insult", upreg = "FT240", downreg = "PBS", flip_log_fold = T) |> add_labels(df = EV_protein_FT240, n_genes = 40, lab_size = 6, x_lim_max = 6, flip_log_fold = T)

VP_OVCAR3

VP_FT240

```

### Fig6.2 GO Plots

#### Load files Proteomics Go Plots


```{r}

source <- paste0(getwd(), "/analysis/proteomics")
PBS_v_OVCAR3_GO_BP <- read.csv(file = paste0(source, "/PBS_v_OVCAR3_GO_BP_OVC_unique.csv"))

```

#### Generate GO plots


```{r fig.width=8, fig.height=8}

plot_GO(data = PBS_v_OVCAR3_GO_BP[PBS_v_OVCAR3_GO_BP$nGenes > 3, ], n_graph = 10) + ggtitle("GO Biological Pathways associated with  \n exo-proteins induced by OVCAR3 EVs") + theme(plot.caption = element_text(hjust = 1, face= "italic"), #Default is hjust=1
        plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot")+ theme(plot.title = element_text(size = 24))



```


## Figure 7

Figure 7 contains boxplots that were generated using R. 


#### FUNCT: subset_dataset()

```{r}
subset_dataset <- function(dataset = NA, subset_by = NA, value_to_select = NA){
  
    # check to see if subsetting requested
  if(is.character(subset_by)){
    # check to see if a value provided. 
    if(!is.list(value_to_select)){
      message("No value provided to subset on.")
    }
    else{
      # in the column subset_by
      # select all of the rows that match "value_to_select"
      dataset = dataset[, pData(dataset)[, subset_by] %in% value_to_select[[1]]] 
    }
  }
  
  return(dataset)
}




multi_subset_test <- function(dataset = NA, subset_pairs = NA){
  if(is.list(subset_pairs)){
    for(name in names(subset_pairs)){
      dataset <- subset_dataset(dataset = dataset, subset_by = name, value_to_select = subset_pairs[name])
    }
  }
  else{
    message("Subset_pairs must be a list.")
  }
  return(dataset)
  
  
}
```

### FUNCT: make_boxplot()


```{r}


make_boxplot <- function(gene = "FOXJ1", point_size = 2, dataset = NA, subset_pairs = NA, x_lab = "EV Treatment", title = "", title_append = "", 
                         add_sig = FALSE, sig_comparisons = c()){
  
  
  # check to see if subsetting requested
  if(is.list(subset_pairs)){
    for(name in names(subset_pairs)){
      dataset <- subset_dataset(dataset = dataset, subset_by = name, value_to_select = subset_pairs[name])
    }
  }
  else{
    message("Subset_pairs must be a list.")
  }
  
  if(nchar(title) == 0){
    title = TeX(paste0("\\textit{", gene, "} ", title_append))
  }
  

  # %in% checks to see if a name is in a list. 
  # I define not in (%ni%) to do the opposite (check if NOT in the list)
  `%ni%` <- Negate(`%in%`)
  gene_not_found <- c()
  if(gene %ni% rownames(assayDataElement(dataset, elt = "q_norm"))){
    return(paste0(gene, " not in list"))
  }
    vector <- as.vector(assayDataElement(dataset[gene, ], elt = "q_norm"))
    plot <- ggplot(
      pData(dataset), 
      aes(x = Treatment,
          y = vector,
          fill = Treatment
          )
      )+
      geom_boxplot(outlier.shape = NA) +
      labs(title = title, 
           y = TeX(paste0("\\textit{", gene, "} Expression")),
           x = x_lab)+
      #facet_wrap(~Patient)+
  
      #geom_jitter(size = point_size, width = 0.25, stroke = 1.3, aes(shape = Patient))+
      
      geom_jitter(size = point_size, width = 0.25, stroke = 1.3)+
  
      #scale_shape_manual(values = c(3, 4, 13, 21, 22, 23, 24))+
  
      theme_bw()+
      
      theme(legend.position="none")+
      
      theme(
        axis.title=element_text(size=20), 
        axis.text=element_text(size=15), 
        plot.title=element_text(size=17))
          
    
    if(add_sig){
      plot <- plot |> add_sig(sig_comparisons)
    }
  
      # scale_x_discrete(labels=c("Fim","Inf","Amp", "Isth"))+
      # scale_fill_manual(values= c("#7CAE00", "#00C1C6", "#F8766D", "#C77CFF"))+
  
      # theme(axis.text.x = element_text(size = 15),
      #       axis.text.y = element_text(size = 12),
      #       axis.title.y = element_text(size = 15),
      #       axis.title.x = element_text(size = 15),
      #       legend.text = element_text(size =12),
      #       legend.title = element_text(size = 12),
      #       strip.text = element_text(size = 15))
    

    return(plot)
}
```

### FUNCT: add_sig()

```{r}
my_comparisons = list(c("PBS", "OVCAR3"), c("FT240", "PBS"), c("OVCAR3", "FT240"))


add_sig <- function(plot, comparisons_wanted = my_comparisons, pnt = 6){
  plot <- plot + 
    stat_compare_means(
      comparisons = comparisons_wanted, 
      method = "t.test", 
      bracket.nudge.y = -0.5,
      size = pnt)+
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.10)))
  return (plot)
}
```



```{r fig.height=5, fig.width=3.7}

# Day 1 Boxplots

gene_list <- c("CCL2", "FLNA", "VCAM1", "TPI1")

day1_sec <- list(Days = c("1-Day"), Segment = c("Secretory"), Treatment = c("OVCAR3", "FT240", "PBS"))


lapply(gene_list, make_boxplot, dataset = EV_q_norm, subset_pairs = day1_sec, title_append = "Day1, Secretory", add_sig = TRUE, sig_comparisons = my_comparisons)


# Day 14 Boxplots Sec

gene_list <- c("TXNIP")

day14_sec <- list(Days = c("14-Day"), Segment = c("Secretory"), Treatment = c("OVCAR3", "FT240", "PBS"))


lapply(gene_list, make_boxplot, dataset = EV_q_norm, subset_pairs = day14_sec, title_append = "Day14, Secretory", add_sig = TRUE, sig_comparisons = my_comparisons)

# Day 14 Boxplots Cil

gene_list <- c("VCAM1")

day14_sec <- list(Days = c("14-Day"), Segment = c("Ciliated"), Treatment = c("OVCAR3", "FT240", "PBS"))


lapply(gene_list, make_boxplot, dataset = EV_q_norm, subset_pairs = day14_sec, title_append = "Day14, Ciliated", add_sig = TRUE, sig_comparisons = my_comparisons)

```




## Supp. Figure 3



```{r}

dataEV_D14 <- EV_q_norm[, pData(EV_q_norm)$Days == "14-Day"]

dataEV_D01 <- EV_q_norm[, pData(EV_q_norm)$Days == "1-Day"]


```



```{r}

t.test_logfold <- function(obj, variable_type = "segment", upreg_group = "Ciliated", downreg_group = "Secretory", elt_name = "q_norm"){
  
  # Get locations of all in the "up" group for that variable
  g_up <- which(pData(obj)[variable_type] == upreg_group)
  # get locations of all in the "down" group for that variable
  g_down <- which(pData(obj)[variable_type] == downreg_group)
  
  # Retrieve normalized expression data from the GeoMxDataSet Object
  obj <-  assayDataElement(object = obj, elt = elt_name)
  
  # For each gene (every row), perform a T-test comparing the columns in the 
  #up_group vs down_group
  
  t.test <- apply(obj, 1, 
              function(x) t.test(
                x[g_up],
                x[g_down]))
  
  #calculate the log_2 Fold Change of up_group compared to down for each row
  
  log_fold <- apply(obj, 1,
                    function(x)  log2(mean(x[g_up])/mean(x[g_down]))
                    )
  
  # convert to data frame
  log_fold <- as.data.frame(log_fold)
  
  # retrieve the p.values from the t.test object and store to a datafram

  df <- do.call(rbind, lapply(t.test, function(x) c(
    p_value = unname(x$p.value))))
  
  # create an adjusted p_value using the Bonferooni correction 

  df <- transform(df, padj = p.adjust(p_value, method = "BH"))
  
  # merge the log_fold values with the p_values
  
  output <- merge(df, log_fold, by = 0)
  
  
  # rename the column rownames to Gene
  
  output <- output |> dplyr::rename(Gene = Row.names)
  #  Add the -log10 of the pvalue for plotting 
  output <- output |> mutate(minuslog10_Pvalue = -log10(p_value))
  
  return(output)
  
}
```



#### 1. Volcano Plot Ciliated vs Secretory

```{r fig.height=5, fig.width=12}

cil_v_sec_t.test_D01 <- t.test_logfold(dataEV_D01, variable_type = "segment", upreg_group = "Ciliated", downreg_group = "Secretory", elt_name = "q_norm")

cil_v_sec_t.test_D01 <- cil_v_sec_t.test_D01 #|> dplyr::rename(c(Marker.name = Gene, p_value_t = p_value, minus_log10.p.value_t = minuslog10_Pvalue))



vp1 <- volcano_plot(df = cil_v_sec_t.test_D01, title_name = "Ciliated vs Secretory Comparison (Day 1)", highlight_genes_purple = c("PAX8"), upreg = "Ciliated", downreg = "Secretory", FC =0.5)



cil_v_sec_t.test_D14 <- t.test_logfold(dataEV_D14, variable_type = "segment", upreg_group = "Ciliated", downreg_group = "Secretory", elt_name = "q_norm")

cil_v_sec_t.test_D14 <- cil_v_sec_t.test_D14 #|> dplyr::rename(c(Marker.name = Gene, p_value_t = p_value, minus_log10.p.value_t = minuslog10_Pvalue))


vp2 <- volcano_plot(df = cil_v_sec_t.test_D14, title_name = "Ciliated vs Secretory Comparison (Day14)", highlight_genes_purple = c("PAX8"), upreg = "Ciliated", downreg = "Secretory", FC =0.5)


vp1 + vp2 


```



### Ciliated v Secretory Marker Confirmation Boxplots


```{r}

Cil_v_Sec_Boxplot_Points <- function(gene = "FOXJ1", point_size = 3, dataset = EV_q_norm){
    set.seed(42)
    vector <- as.vector(assayDataElement(dataset[gene, ], elt = "q_norm"))
    plot <- ggplot(pData(dataset),
      aes(x = segment,
          y = vector,
          fill = segment
          )
      )+
      geom_boxplot(outlier.shape = NA)+
      labs(title = TeX(paste0("\\textit{", gene, "}")),
           y = TeX(paste0("\\textit{", gene, "} Expression")),
           x = "Cell Type")+

      geom_jitter(size = point_size, width = 0.25, stroke = 0.3)+

      # geom_jitter(size = point_size, width = 0.25, stroke = 1.3,
      #             aes(shape = patient))+

      # scale_shape_manual(values = c(21, 22, 23))+
      scale_fill_manual(values =c("green", "red"))+



      theme_bw()+
      #
      # theme(axis.text.x = element_text(size = 15),
      #       axis.text.y = element_text(size = 12),
      #       axis.title.y = element_text(size = 15),
      #       axis.title.x = element_text(size = 15),
      #       legend.text = element_text(size =12),
      #       legend.title = element_text(size = 12),
      #       strip.text = element_text(size = 15))

    
    stat_compare_means(
      comparisons = list(c("Ciliated", "Secretory")),
      method = "t.test", 
      bracket.nudge.y = -0.5,
      size = point_size+5)+
      
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.10)))
    
    
  return(plot)

}

```




```{r}



A <- Cil_v_Sec_Boxplot_Points(gene = "FOXJ1", dataset = dataEV_D01)

B <- Cil_v_Sec_Boxplot_Points(gene = "PAX8", dataset = dataEV_D01)


C <- Cil_v_Sec_Boxplot_Points(gene = "FOXJ1", dataset = dataEV_D14)

D <- Cil_v_Sec_Boxplot_Points(gene = "PAX8", dataset = dataEV_D14)


A+B #day1

C+D

```


Not included, Supplemental Figures 4-9


## Supplemental Figure 10

OVCAR3 v PBS (sec, day 1)
FT240 v PBS (sec, day 1)
OVCAR3 v PBS (cil, day 1)
FT240 v PBS (cil, day 1)

```{r fig.height=8, fig.width=10}

genes_in_pathway <- c("NFKB1", "NFKB2", "CXCL10", "CXCL11", "IDO1", "IRF8", "IL6", "STAT1", "HLA-A", "HIF1A", "HLA-DRA", "IRF9")



# Day 1 Volcano Plots


# choose label size
size <- 3
genes <- 0



d1_sec_plot_OVC <- volcano_plot(df = sec_D1_PBS.vs.OVCAR3 , title = "OVCAR3 vs PBS \n(Secretory Cells, Day 1)", 
                                highlight_genes_purple= genes_in_pathway,
                                upreg = "OVCAR3", downreg = "PBS", flip_log_fold = F, lab_size = size, num_genes = genes)+ theme(
  axis.title = element_text(size = 20)) 


d1_cil_plot_OVC <- volcano_plot(df = cil_D1_PBS.vs.OVCAR3 , title = "OVCAR3 vs PBS \n(Ciliated Cells, Day 1)", 
                                highlight_genes_purple= genes_in_pathway,
                                upreg = "OVCAR3", downreg = "PBS", flip_log_fold = F, lab_size = size, num_genes = genes) + theme(
  axis.title = element_text(size = 20))





d1_sec_plot_FT <- volcano_plot(df = sec_D1_PBS.vs.FT240, title = "FT240 vs PBS \n(Secretory Cells, Day 1)", 
                                highlight_genes_purple= genes_in_pathway,
                               upreg = "FT240", downreg = "PBS", flip_log_fold = F, lab_size = size, num_genes = genes)+ theme(
  axis.title = element_text(size = 20)) 


d1_cil_plot_FT <- volcano_plot(df = cil_D1_PBS.vs.FT240 , title = "FT240 vs PBS \n(Ciliated Cells, Day 1)", 
                               highlight_genes_purple= genes_in_pathway,
                               upreg = "FT240", downreg = "PBS", flip_log_fold = F, lab_size = size, num_genes = genes) + theme(
  axis.title = element_text(size = 20))



  (d1_sec_plot_OVC | d1_sec_plot_FT) / 
  (d1_cil_plot_OVC | d1_cil_plot_FT) 

```




## Supplemental Figure 11

Supplemental Figure 11 is not included in this document, because it was created using data from another paper.

To see Supp. Fig 11 code, go to the document HGSOC_precursor_comparison.Rmd. 




## Supplemental Figure 12-13

### Load FT240 Treatment GO Files

files to load:

Source: /corrected_GO_analysis
Names:
1. cil_downreg_FT240.xlsx
2. cil_upreg_FT240.xlsx
3. sec_upreg_FT240.xlsx
4. sec_downreg_FT240.xlsx


```{r}


#####---CORRECTED GO ANALYSIS---#####


# source of all related .xlsx files
source <- paste0(getwd(), "/corrected_GO_analysis/")

# Read in the .xlsx files to be used.

cell_types <- c("cil", "sec")
directions <- c("upreg", "downreg")
analyses <- c("GO_BP", "GO_CC", "GO_MF")


# initialize list

FT240_GO <- list()

rename_vec <- c("Fold Enrichment" = "Fold.Enrichment",
                "Enrichment FDR" = "Enrichment.FDR")


for(cell in cell_types){
  
  FT240_GO[[cell]] <- list()
  
  for(dir in directions){
    
    FT240_GO[[cell]][[dir]] <- list()
    
    file_name <- paste0("/", cell, "_", dir, "_FT240.xlsx")
    
    for(analysis in analyses){
      
      FT240_GO[[cell]][[dir]][[analysis]] <- read_xlsx(path = paste0(source, file_name), sheet = analysis) |> 
       rename_with(
        ~ case_when(
        . == "Fold Enrichment" ~ "Fold.Enrichment",
        . == "Enrichment FDR" ~ "Enrichment.FDR",
        TRUE ~ .))
    }
  }
}


# Note on column renaming
# ~ creates an anonymous function
# . represents each column name
# so we check to see if column name matches desired, then rename to desired.
# TRUE ~ . means that all the other column names are unchanged




```


### Supp Figure 12

General Outline:

GO_BP_sec_FT240_up + GO_MF_sec_FT240_up + GO_CC_sect_FT240_up

GO_BP_cil_FT240_up + GO_MF_cil_FT240_up + GO_CC_cil_FT240_up

```{r fig.height=21, fig.width=24}

# up in secretory cells 

graph_A <- plot_GO(data = FT240_GO$sec$upreg$GO_BP) + ggtitle("GO Biological Pathway Upregulated in \n Secretory Cells by 1 Day \n FT240 EV Treatment")

graph_B <- plot_GO(data = FT240_GO$sec$upreg$GO_MF) + ggtitle("GO Molecular Function Upregulated in \n Secretory Cells by 1 Day \n FT240 EV Treatment")


graph_C <- plot_GO(data = FT240_GO$sec$upreg$GO_CC) + ggtitle("GO Cellular Components Upregulated in \n Secretory Cells by 1 Day \n FT240 EV Treatment")


# up in ciliated cells

graph_D <- plot_GO(data = FT240_GO$cil$upreg$GO_BP) + ggtitle("GO Biological Pathway Upregulated in \n Ciliated Cells by 1 Day \n FT240 EV Treatment")

graph_E <- plot_GO(data = FT240_GO$cil$upreg$GO_MF) + ggtitle("GO Molecular Function Upregulated in \n Ciliated Cells by 1 Day \n FT240 EV Treatment")

graph_F <- plot_GO(data = FT240_GO$cil$upreg$GO_CC) + ggtitle("GO Cellular Components Upregulated in \n Ciliated Cells by 1 Day \n FT240 EV Treatment")


graph_A + graph_B + graph_C + graph_D + graph_E + graph_F

```


### Supp Figure 13
General Outline:

GO_BP_sec_FT240_down + GO_MF_sec_FT240_down + GO_CC_sect_FT240_down

GO_BP_cil_FT240_down + GO_MF_cil_FT240_down + GO_CC_cil_FT240_down

```{r fig.height=21, fig.width=24}

# up in secretory cells 

graph_A <- plot_GO(data = FT240_GO$sec$downreg$GO_BP) + ggtitle("GO Biological Pathway Downregulated in \n Secretory Cells by 1 Day \n FT240 EV Treatment")

graph_B <- textGrob('No GO Molecular Function Pathways \n with at least 5 genes \n downregulated in secretory cells \n by FT240 EV treatment', gp=gpar(fontsize=25))

#plot_GO(data = FT240_GO$sec$downreg$GO_MF) + ggtitle("GO Molecular Function Downregulated in \n Secretory Cells by 1 Day \n FT240 EV Treatment")


graph_C <- plot_GO(data = FT240_GO$sec$downreg$GO_CC) + ggtitle("GO Cellular Components Downregulated in \n Secretory Cells by 1 Day \n FT240 EV Treatment")


# up in ciliated cells

graph_D <- plot_GO(data = FT240_GO$cil$downreg$GO_BP) + ggtitle("GO Biological Pathway Downregulated in \n Ciliated Cells by 1 Day \n FT240 EV Treatment")

graph_E <- plot_GO(data = FT240_GO$cil$downreg$GO_MF) + ggtitle("GO Molecular Function Downregulated in \n Ciliated Cells by 1 Day \n FT240 EV Treatment")

graph_F <- plot_GO(data = FT240_GO$cil$downreg$GO_CC) + ggtitle("GO Cellular Components Downregulated in \n Ciliated Cells by 1 Day \n FT240 EV Treatment")


graph_A + graph_B + graph_C + graph_D + graph_E + graph_F

```



## Supplemental Figure 14-15


### Load Files from Mihaela

```{r}
kegg_and_reactome_files <- paste0(getwd(), "/corrected_GO_analysis/kegg_and_reactome/files_for_plotting")


pathway_files <- list()

days <- c("Day1")
cell_types <- c("Sec", "Cil")
comparisons <- c("PBS.VS.OVCAR3", "PBS.VS.FT240")
categories <- c("KEGG.UP", "KEGG.DOWN", "RCT.UP", "RCT.DOWN")


# Initialize the main list
pathway_files <- list()


for(day in days){
  pathway_files[[day]] <- list()
  
  day_short <- switch(day,
                      Day1 = "1Day",
                      # default value if no match
                      NA_character_)
  
  for(cell in cell_types){
    pathway_files[[day]][[cell]] <- list()
    
    # define a longer name based on the cell type used
    long_name <- switch(cell,
                        Sec = "Secretory",
                        Cil = "Ciliated",
                        # default value if no match
                        NA_character_)
    # get the name of the file to access
    file_name <- paste(day_short, long_name, "ttest", sep ="_") 
      
    for(comp in comparisons) {
      pathway_files[[day]][[cell]][[comp]] <- list()
      
      for(cat in categories){
        
        # define the sheet name
        sheet_name <- paste(comp, cat, sep = ".")
        
        # Read in and assign file to appropriate location
        pathway_files[[day]][[cell]][[comp]][[cat]] <- read_xlsx(
          path = paste0(kegg_and_reactome_files, "/", file_name, ".xlsx"),
          sheet = sheet_name,
          range = cell_cols("B:K")
        )
        
      }
    }
  }
  
  
}

# Test

pathway_files$Day1$Sec$PBS.VS.OVCAR3$KEGG.UP




```


```{r}

path_plot <- function(data = pathway_files$Day1$Cil$PBS.v.OVCAR3$RCT.DOWN, n_graph = 10, optional_title = "No significantly enriched genes"){
  
  data <- data[data$nGenes > 4, ]
  plot1_data <- head(data, n = n_graph) 
  

  
  plot1_data <- plot1_data |> mutate(minuslog10pvalue = -log10(`p-value`))
  
  #plot1_data <- plot1_data[plot1_data$nGenes > 2 & plot1_data$`p-value` < 0.05, ]
  
  if(nrow(plot1_data) == 0){
    
    plot <- textGrob(label = optional_title, gp = gpar(fontsize = 35))
    return(plot)
  }
  
  
  
  # Define a list of abbreviations
  abbreviations <- c("response" = "resp", "mediated by" = "med", 
                   "chemotaxis" = "chemotx", "migration" = "migr", 
                   "immune" = "imm", "antimicrobial" = "antimicro",
                   "Homo" = "", "sapiens" = "", "\\(human\\)" = "",
                   "-" = "")

    # Function to replace words with abbreviations
  simplify_names <- function(name) {
      for (word in names(abbreviations)) {
        name <- gsub(word, abbreviations[[word]], name)
      }
      return(name)
    }

  # Apply the function to the description column
  plot1_data$pathway <- sapply(plot1_data$pathway, simplify_names)
    


  
  plot_df <- plot1_data  |> dplyr::arrange(minuslog10pvalue) |> mutate(pathway = factor(pathway, levels = pathway))



    
    
    test<-   ggplot(data = plot_df,  aes(x=minuslog10pvalue, y= pathway)) +
        geom_segment(aes(x= 0, xend=minuslog10pvalue), linewidth = 1) +
        geom_point(aes(size= nGenes)) +
        theme_bw() +
        scale_y_discrete(labels = function(x) str_wrap(x, width = 25))+
      labs(x = "-log10(p-value)")
    
    
    plot <- test + theme(
      axis.title=element_text(size=24),
      axis.text=element_text(size=18),
      plot.title=element_text(size=18),
      legend.text =element_text(size=18),
      legend.title = element_text(size = 18),
      axis.text.y = element_text(vjust = 0.5))
    
    return(plot)
}



```

### Supp Fig 14

```{r fig.height=21, fig.width=36}

library(patchwork)

# Up in Sec

sec_data <- pathway_files$Day1$Sec$PBS.VS.OVCAR3
cil_data <- pathway_files$Day1$Cil$PBS.VS.OVCAR3

graph_A <- path_plot(data = sec_data$RCT.UP, n_graph = 10) + ggtitle("Reactome Upregulated in Secretory Cells \n by 1 Day OVCAR3 EV Treatment")

graph_B <- path_plot(data = sec_data$KEGG.UP, n_graph = 10) + ggtitle("Kegg Upregulated in Secretory Cells \n by 1 Day OVCAR3 EV Treatment")

# Up in Cil
graph_C <- path_plot(data = cil_data$RCT.UP, n_graph = 10) + ggtitle("Reactome Upregulated in Ciliated Cells \n by 1 Day OVCAR3 EV Treatment")

graph_D <- path_plot(data = cil_data$KEGG.UP, n_graph = 10) + ggtitle("Kegg Upregulated in Ciliated Cells \n by 1 Day OVCAR3 EV Treatment")


# Down in Sec
graph_E <- path_plot(data = sec_data$RCT.DOWN, n_graph = 10) + ggtitle("Reactome Downregulated in Secretory Cells \n by 1 Day OVCAR3 EV Treatment")

graph_F <- path_plot(data = sec_data$KEGG.DOWN, n_graph = 10) + ggtitle("Kegg Downregulated in Secretory Cells \n by 1 Day OVCAR3 EV Treatment")

# Down in Cil
graph_G <- path_plot(data = cil_data$RCT.DOWN, n_graph = 10, 
                     optional_title = "No Reactome Pathways \n Downregulated by OVCAR3 EVs \n in Ciliated Cells.") 
                      #+ ggtitle("Reactome Downregulated in Ciliated Cells \n by 1 Day OVCAR3 EV Treatment")

graph_H <- path_plot(data = cil_data$KEGG.DOWN, n_graph = 10) + ggtitle("Kegg Downregulated in Ciliated Cells \n by 1 Day OVCAR3 EV Treatment")



graph_A + graph_B + graph_C + graph_D + graph_E + graph_F + graph_G + graph_H +  plot_layout(nrow=2)


```

```{r fig.height=21, fig.width=36}
sec_data <- pathway_files$Day1$Sec$PBS.VS.FT240
cil_data <- pathway_files$Day1$Cil$PBS.VS.FT240

graph_A <- path_plot(data = sec_data$RCT.UP, n_graph = 10) + ggtitle("Reactome Upregulated in Secretory Cells \n by 1 Day FT240 EV Treatment")

graph_B <- path_plot(data = sec_data$KEGG.UP, n_graph = 10) + ggtitle("Kegg Upregulated in Secretory Cells \n by 1 Day FT240 EV Treatment")

# Up in Cil
graph_C <- path_plot(data = cil_data$RCT.UP, n_graph = 10) + ggtitle("Reactome Upregulated in Ciliated Cells \n by 1 Day FT240 EV Treatment")

graph_D <- path_plot(data = cil_data$KEGG.UP, n_graph = 10) + ggtitle("Kegg Upregulated in Ciliated Cells \n by 1 Day FT240 EV Treatment")


# Down in Sec
graph_E <- path_plot(data = sec_data$RCT.DOWN, n_graph = 10,
                     optional_title = "No Reactome Pathways \n Downregulated by FT240 EVs \n in Secretory Cells."
                       ) + ggtitle("Reactome Downregulated in Secretory Cells \n by 1 Day FT240 EV Treatment")

graph_F <- path_plot(data = sec_data$KEGG.DOWN, n_graph = 10,
                     optional_title = "No Kegg Pathways \n Downregulated by FT240 EVs \n in Secretory Cells."
                     ) #+ ggtitle("Kegg Downregulated in Secretory Cells \n by 1 Day OVCAR3 EV Treatment")

# Down in Cil
graph_G <- path_plot(data = cil_data$RCT.DOWN, n_graph = 10, 
                     optional_title = "No Reactome Pathways \n Downregulated by FT240 EVs \n in Ciliated Cells.") +
          ggtitle("Reactome Downregulated in Ciliated Cells \n by 1 Day OVCAR3 EV Treatment")


graph_H <- path_plot(data = cil_data$KEGG.DOWN, n_graph = 10,
                     optional_title = "No Kegg Pathways \n Downregulated by FT240 EVs \n in Ciliated Cells.") 
                     #+ ggtitle("Kegg Downregulated in Ciliated Cells \n by 1 Day OVCAR3 EV Treatment")



graph_A + graph_B + graph_C + graph_D + graph_E + graph_F + graph_G + graph_H +  plot_layout(nrow=2)
```



# Supp. Fig. 16


```{r fig.height=15, fig.width=18}


create_heatmap_unordered <- function(dataset = EV_1_day_sec, 
                                  order_treatment = c("PBS", "FT240", "OVCAR3"), 
                                  order_Pts = pt_order,
                                  annotation_colors = my_colors, 
                                  title = "Insert Title Here",
                                  legend = TRUE, 
                                  font = 20,
                                  gene_list = GOI_new,
                                  split_point = 36){
  
  # create the annotation column
  annotation_col <- create_annotation_col(dataset = dataset, order_treatment = order_treatment, order_Pts = order_Pts)
  # create the matrix with expression data
  matrix_reordered <- create_ordered_matrix(dataset = dataset, order_treatment = order_treatment, order_Pts = order_Pts, gene_list = gene_list)
  
  
  # First, convert your matrix to a data frame for easier manipulation
  mat_df <- as.data.frame(matrix_reordered)

  # Transpose the matrix so that columns become rows (for easier merging with annotation)
  mat_df <- t(mat_df)
  
  

  # Merge the matrix data with the annotation data based on column names
  merged_df <- merge(mat_df, annotation_col, by = "row.names" )
  
  # Now, group by treatment and patient and compute the row-wise average
  averaged_df <- merged_df %>%
    group_by(Treatment, Patient) %>%
    summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))  # Average each row, excluding non-numeric columns
  
  
  # create new annotation df
  
  annotation_df <- as.data.frame(averaged_df[1:2])
  
  matrix_averaged <- t(as.matrix(averaged_df[3: ncol(averaged_df)]))
  
  colnames(matrix_averaged) <- rownames(annotation_df)
  
  
  # implement row scaling - z-score normalization across row
  matrix_scaled <- t(scale(t(matrix_averaged)))
  
  
  # create complext heatmap object 

  heatmap = Heatmap(
    matrix_scaled,
    name = title,
    col = colorRamp2(
    c(-3.5, 0, 3.5),  # Set color at 0 to white
    c("navy", "white", "firebrick3")),
  
  # colorRampPalette(c("navy", "white", "firebrick3"))(21),
    cluster_rows = TRUE,
    cluster_columns = TRUE,

  
  
    top_annotation = HeatmapAnnotation(df = annotation_df, col = annotation_colors, 
                                       annotation_name_gp = gpar(fontsize = font),
                                       show_annotation_name = legend, # turn off annotation name if legend = F
                                       annotation_legend_param = list(
                                         ncol = 4, # Set the number of columns for the legend
                                         gp = gpar(fontsize = font),
                                         title_gp = gpar(fontsize = font + 2), # title font size
                                         labels_gp = gpar(fontsize = font)   # Label font size  
      )),  # Top annotations
    column_title = title, # add a title to the heatmap
    show_heatmap_legend = legend,
    show_row_names = TRUE,
    show_column_names = FALSE,
    border = TRUE,
    row_gap = unit(1, "mm"),  # Customize row gaps
    heatmap_legend_param = list(
      title = "Z-score", legend_direction = "horizontal", side = "right", title_gp = gpar(fontsize = font+2),  # Title font size
    labels_gp = gpar(fontsize = font)   # Label font size
    ),
  
    row_names_gp = gpar(fontsize = font-8),
    row_title_gp = gpar(fontsize = font+8),   # Row title font size
    column_title_gp = gpar(fontsize = font+8) # Column title font size
  )


  
  return(heatmap)
}





heatmap1 = create_heatmap_unordered(dataset = EV_1_day_sec, 
                  order_treatment = c("PBS", "FT240", "OVCAR3"), 
                  order_Pts = pt_order, 
                  annotation_colors = my_colors,
                  title = "1-Day Treatment, \n Secretory Cells", legend = FALSE, font = 24,
                  gene_list = GOI_sec,
                  split_point = 36)


heatmap2 = create_heatmap_unordered(dataset = EV_1_day_cil, 
                      order_treatment = c("PBS", "FT240", "OVCAR3"), 
                      order_Pts = pt_order, 
                      annotation_colors = my_colors,
                      title = "1-Day Treatment, \n Ciliated Cells", font = 24,
                      gene_list = GOI_cil,
                      split_point = 13)

heatmap1 

heatmap2


```



```{r}

sessionInfo()
```

